---
title: "Plotting differential expression results"
bibliography: bibliography.bib
output:
   BiocStyle::html_document:
      code_folding: hide
      number_sections: yes
      toc: yes  
---

This notebook prints volcano plots of differential expression (DE) results obtained using
a mixed modelling approach (@nebula), and testing test for DE between tdTom+ and tdTom- 
cells in the knockout chimeras, contrasted with the same change in the WT chimeras


# Set-up and loading DE results


```{r, echo=FALSE, output=FALSE, message=FALSE}

knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    cache=FALSE#,
    #dev = c("pdf"),
    #dpi=300
)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
```


# Overview plots

```{r}
all_Mixl1 <- read.table("markers_DE_celltype/Mixl1_DE.csv",sep=",",header=TRUE)
all_T <- read.table("markers_DE_celltype/T_DE.csv",sep=",",header=TRUE)
table(all_Mixl1$cell_type)
table(all_T$cell_type)
```


## Plotting number of DE genes

We plot the number of significantly DE genes across different cell types. Note that significance depends on cell numbers, and therefore smaller cell types are less likely to have significantly DE genes. 

## For the T chimeras

```{r nr_DE_T,fig.height=4,fig.width=4}
nr_DE_genes_T <- as.data.frame(table(all_T$cell_type))
nr_DE_genes_T_top <- nr_DE_genes_T[order(nr_DE_genes_T$Freq,decreasing=T)[1:8],]
nr_DE_genes_T_top$Var1 <- factor(nr_DE_genes_T_top$Var1,levels=nr_DE_genes_T_top$Var1)
ggplot(nr_DE_genes_T_top,aes(x=Var1,y=Freq)) + geom_bar(stat="identity") + theme_classic()+
  xlab("") + ylab("nr. of DE genes") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## For the Mixl1 chimeras

```{r nr_DE_Mixl1,fig.height=4,fig.width=4}
nr_DE_genes_Mixl1 <- as.data.frame(table(all_Mixl1$cell_type))
nr_DE_genes_Mixl1_Top <- nr_DE_genes_Mixl1[order(nr_DE_genes_Mixl1$Freq,decreasing=T)[1:8],]
nr_DE_genes_Mixl1_Top$Var1 <- factor(nr_DE_genes_Mixl1_Top$Var1,levels=nr_DE_genes_Mixl1_Top$Var1)
ggplot(nr_DE_genes_Mixl1_Top,aes(x=Var1,y=Freq)) + geom_bar(stat="identity") + theme_classic()+
  xlab("") + ylab("nr. of DE genes") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Volcano plots 

For a number of specific cell types, we illustrate the DE genes using volcano plots. 

```{r}
library(ggplot2)
library(ggrepel)
imprinted_genes <- read.table("list_imprinted_genes_Santini_et_al.csv",sep=",")$V1
volcano_plot <- function(vector_FDR,vector_effect_sizes,thresh_FDR=0.1,max_highlight=50,
                         x_label="coefficient",title=""){
    df <- data.frame(FDR=vector_FDR,effect=vector_effect_sizes,gene=names(vector_FDR))
    df <- df[complete.cases(df),]
    rownames(df) <- df$gene
    df$sig <- "not significant"
    df$sig[df$FDR<thresh_FDR] <- "significant"
    max_highlight <- min(sum(df$FDR<0.1),max_highlight)
    highlight <- df$FDR < thresh_FDR &((df$effect > 0 & df$effect >= sort(df$effect[df$FDR<thresh_FDR],decreasing=TRUE)[max_highlight])|(df$effect < 0 & df$effect <= sort(df$effect[df$FDR <thresh_FDR],decreasing=FALSE)[max_highlight]))
    highlight[highlight & (df$effect > 0)] <- "up"
    highlight[highlight==TRUE & (df$effect < 0)] <- "down"
    df$highlight <- highlight
    df$highlight[df$gene%in%imprinted_genes & df$highlight!="FALSE"] <- "imprinted"
    df$sig[df$sig=="significant" & df$effect < 0] <- "down"
    df$sig[df$sig=="significant" & df$effect > 0] <- "up"
    df$sig[df$highlight == "imprinted"] <- "imprinted"
    pp <- ggplot(df, aes(x=effect, y=-log10(FDR),color=sig)) +
        geom_point(size=5, alpha=0.1) +
        ggtitle(title) + scale_color_manual(values=c("FALSE"="grey","up"="darkblue","down"="darkred","imprinted"="purple"))+
        labs(y=expression('-Log'[10]*' FDR'), x=x_label) +
        theme_classic(base_size=12) +
        theme(legend.position="none", plot.title = element_text(size = rel(1), hjust = 0.5))+
        geom_text_repel(data=df[!(highlight==FALSE),],
                        aes(x = effect, y = -log10(FDR),label=gene),max.overlaps=100,
                        segment.color = 'transparent')+
        scale_y_continuous(trans=scales::pseudo_log_trans(sigma=1.1,base = 1.7))+
        scale_x_continuous(trans=scales::pseudo_log_trans(sigma=1.1,base = 1.1))+
        geom_hline(yintercept = 1) + xlim(c(-0.3+min(df$effect[df$highlight=="down"]),0.3+max(df$effect[df$highlight=="up"])))
    return(pp)

}

wrapper_volcano_plot <- function(path_to_file_name, target, cell_type){
  res <- read.table(path_to_file_name,sep=",",header=TRUE)
  # occassionally a duplicated name resulting from conversion to gene names, e.g.Dohh
  res$DE_gene_name[duplicated(res$DE_gene_name)] <- paste0(res$DE_gene_name[duplicated(res$DE_gene_name)],"_1")
  vector_FDR <- res$FDR
  vector_effect_sizes=res$logFC_target_tomatoTRUE
  names(vector_FDR) <- res$DE_gene_name
  names(vector_effect_sizes) <- res$DE_gene_name
  print(volcano_plot(vector_FDR,vector_effect_sizes,thresh_FDR=0.1,max_highlight=22,
                           x_label="logFC",title=paste(target," - ",cell_type)))
}
```

## T

```{r T_LPM, fig.width=5,fig.height=4}
wrapper_volcano_plot("markers_DE_celltype/Lateral_plate_mesoderm_T_DE.csv","T","LPM")
```

```{r T_LM, fig.width=5,fig.height=3}
wrapper_volcano_plot("markers_DE_celltype/Limb_mesoderm_T_DE.csv","T","Limb mesoderm")
```
```{r T_NMPs_mesoderm_biased, fig.width=7,fig.height=4.5}
wrapper_volcano_plot("markers_DE_celltype/NMPs_Mesoderm-biased_T_DE.csv","T","NMPs mesoderm biased")
```
```{r T_NMPs, fig.width=5,fig.height=3}
wrapper_volcano_plot("markers_DE_celltype/NMPs_T_DE.csv","T","NMPs")
```
```{r T_intermdiate_mesoderm, fig.width=5,fig.height=3}
wrapper_volcano_plot("markers_DE_celltype/Intermediate_mesoderm_T_DE.csv","T","Intermediate mesoderm")
```

## Mixl1

```{r Mixl1_LPM, fig.width=5,fig.height=3}
wrapper_volcano_plot("markers_DE_celltype/Lateral_plate_mesoderm_Mixl1_DE.csv","Mixl1","LPM")
```
```{r Mixl1_Mesenchyme, fig.width=8,fig.height=6}
wrapper_volcano_plot("markers_DE_celltype/Mesenchyme_Mixl1_DE.csv","Mixl1","Mesenchyme")
```
```{r Mixl1_Blood_progenitors, fig.width=7,fig.height=5}
wrapper_volcano_plot("markers_DE_celltype/Blood_progenitors_Mixl1_DE.csv","Mixl1","Blood_progenitors")
```

```{r Mixl1_Ery, fig.width=10,fig.height=7}
wrapper_volcano_plot("markers_DE_celltype/Erythroid_Mixl1_DE.csv","Mixl1","Erythroid")
```

```{r Mixl1_PST, fig.width=6,fig.height=4}
wrapper_volcano_plot("markers_DE_celltype/Posterior_somitic_tissues_Mixl1_DE.csv","Mixl1","Posterior_somitic_tissues")
```

# Session Info

```{r}
sessionInfo()
```

# References