---
title: "Comparison of T and Mixl1 chimeras"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes
---

```{r, echo=FALSE, output=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = "pdf",
    dpi=300
)

library(MAYA)

source("core_functions_dynamic.R")
source("chimera_core_functions_big_atlas.R")

Celltype <- "Limb_mesoderm"

correlation_pseudotime_Mixl1 <- readRDS("results_no_split/correlation_pseudotime_Mixl1.rds")
correlation_pseudotime_T <- readRDS("results_no_split/correlation_pseudotime_T.rds")
correlation_pseudotime_WT <- readRDS("results_no_split/correlation_pseudotime_WT.rds")

Celltypes <- read.table("lineage_names.txt")$V1
names(correlation_pseudotime_Mixl1) <- Celltypes
names(correlation_pseudotime_T) <- Celltypes
names(correlation_pseudotime_WT) <- Celltypes

chimeraT <- readRDS("../results/chimeraT_2020_extended_mapping_rev.rds")
chimeraT <- logNormCounts(chimeraT,size_factors=chimeraT$sizeFactor)
chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraWT <- logNormCounts(chimeraWT,size_factors=chimeraWT$sizeFactor)
chimeraMixl1 <- readRDS("../results/chimeraMixl1_2020_extended_mapping_rev.rds")

cor_T <- readRDS(paste0("results_no_split/",Celltype,"/",Celltype,"_cor_cells_T.rds"))
cor_WT <- readRDS(paste0("results_no_split/",Celltype,"/",Celltype,"_cor_cells_WT.rds"))
cor_Mixl1 <- readRDS(paste0("results_no_split/",Celltype,"/",Celltype,"_cor_cells_Mixl1.rds"))
atlas_meta <- readRDS("../fromIvan_new_atlas/integrated_meta_celltype_clus.rds")
chimeraT <- refine_celltype_stage(chimeraT,cor_T)
chimeraWT <- refine_celltype_stage(chimeraWT,cor_WT)
chimeraMixl1 <- refine_celltype_stage(chimeraMixl1,cor_Mixl1)
```

# Expression of T/Mixl1 in Mixl1/T chimeras

```{r}
chimeraMixl1 <- logNormCounts(chimeraMixl1)
chimeraT <- logNormCounts(chimeraT)
chimeraWT <- logNormCounts(chimeraWT)
chimeraMixl1$T_expr <- as.vector(logcounts(chimeraMixl1["ENSMUSG00000062327",]))
chimeraT$Mixl1_expr <- as.vector(logcounts(chimeraT["ENSMUSG00000026497",]))
chimeraWT$T_expr <- as.vector(logcounts(chimeraWT["ENSMUSG00000062327",]))
chimeraWT$Mixl1_expr <- as.vector(logcounts(chimeraWT["ENSMUSG00000026497",]))
```


We plot T expression for WT chimeras for those cell types for which both tomato positive and tomato negative cells exist and either tomato positive or tomato negative cells have an average expression of at least 0.033, and for which there are at least 50 tomato positive cells.

```{r Mixl1_expr_WT_chimeras}
chimeraWT_overall_Mixl1_expression_celltype <- as.data.frame(colData(chimeraWT)[c("Mixl1_expr","celltype.mapped","tomato")]) %>% dplyr::group_by_all()
overall_number_tomato_pos <- table(chimeraWT_overall_Mixl1_expression_celltype$celltype.mapped[chimeraWT_overall_Mixl1_expression_celltype$tomato])
celltypes.keep <- names(overall_number_tomato_pos)[overall_number_tomato_pos>=50]
chimeraWT_overall_Mixl1_expression_celltype <- chimeraWT_overall_Mixl1_expression_celltype[chimeraWT_overall_Mixl1_expression_celltype$celltype.mapped%in%celltypes.keep,]
chimeraWT_overall_Mixl1_expression_celltype_count <- aggregate(Mixl1_expr~.,chimeraWT_overall_Mixl1_expression_celltype,mean)
celltypes_filter_1 <- names(table(chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped))[table(chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped) == 2]
chimeraWT_overall_Mixl1_expression_celltype_count <- chimeraWT_overall_Mixl1_expression_celltype_count[chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped%in%celltypes_filter_1,]
celltypes_filter <- sapply(1:length(chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped),
                           function(x) any(chimeraWT_overall_Mixl1_expression_celltype_count$Mixl1_expr[chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped==chimeraWT_overall_Mixl1_expression_celltype_count$celltype.mapped[x]] >= 0.033))
ggplot(chimeraWT_overall_Mixl1_expression_celltype_count[celltypes_filter,],aes(x=celltype.mapped,y=Mixl1_expr,fill=tomato)) + geom_bar(stat="identity",position = "dodge") + theme_classic()+coord_flip() + scale_fill_manual(values=c("TRUE"="darkred","FALSE"="darkblue"),labels=c("positive","negative")) + ggtitle("Average Mixl1 expression per cell type\nin WT chimeras")+scale_y_continuous(trans=pseudo_log_trans(sigma=0.05,base=500),expand = c(0, 0),breaks=c(0.1,1,10,100))+xlab("")
```

We plot T expression for Mixl1 chimeras for those cell types for which both tomato positive and tomato negative cells exist and either tomato positive or tomato negative cells have an average expression of at least 0.033, and for which there are at least 50 tomato positive cells. 

```{r Mixl1_expr_T_chimeras}
chimeraT_overall_Mixl1_expression_celltype <- as.data.frame(colData(chimeraT)[c("Mixl1_expr","celltype.mapped","tomato")]) %>% dplyr::group_by_all()
overall_number_tomato_pos <- table(chimeraT_overall_Mixl1_expression_celltype$celltype.mapped[chimeraT_overall_Mixl1_expression_celltype$tomato])
celltypes.keep <- names(overall_number_tomato_pos)[overall_number_tomato_pos>=50]
chimeraT_overall_Mixl1_expression_celltype <- chimeraT_overall_Mixl1_expression_celltype[chimeraT_overall_Mixl1_expression_celltype$celltype.mapped%in%celltypes.keep,]
chimeraT_overall_Mixl1_expression_celltype_count <- aggregate(Mixl1_expr~.,chimeraT_overall_Mixl1_expression_celltype,mean)
celltypes_filter_1 <- names(table(chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped))[table(chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped) == 2]
chimeraT_overall_Mixl1_expression_celltype_count <- chimeraT_overall_Mixl1_expression_celltype_count[chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped%in%celltypes_filter_1,]
celltypes_filter <- sapply(1:length(chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped),
                           function(x) any(chimeraT_overall_Mixl1_expression_celltype_count$Mixl1_expr[chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped==chimeraT_overall_Mixl1_expression_celltype_count$celltype.mapped[x]] >= 0.033))
ggplot(chimeraT_overall_Mixl1_expression_celltype_count[celltypes_filter,],aes(x=celltype.mapped,y=Mixl1_expr,fill=tomato)) + geom_bar(stat="identity",position = "dodge") + theme_classic()+coord_flip() + scale_fill_manual(values=c("TRUE"="darkred","FALSE"="darkblue"),labels=c("positive","negative")) + ggtitle("Average Mixl1 expression per cell type\nin T chimeras") +scale_y_continuous(trans=pseudo_log_trans(sigma=0.05,base=500),expand = c(0, 0),breaks=c(0.01,0.1,0.2,0.4))+xlab("")
```


