---
title: "perturbSuite_DE_dynamic"
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes  
---

This notebook performs perturbSuite_DE_dynamic for specified chimera (T or Mixl1) and specified lineage. 

# Set-up and loading data and intermediate results

```{r, echo=FALSE, output=FALSE, message=FALSE}
target <- params[2]
lineage <- params[3]

source("core_functions_dynamic.R")
source("core_functions_DA_testing_general.R")
source("chimera_core_functions_big_atlas.R")
library(ggplot2)
library(ggrepel)

chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraTarget <- readRDS(paste0("../results/chimera",target,"_2020_extended_mapping_rev.rds"))
correlation_pseudotime_WT <- readRDS("results_no_split/correlation_pseudotime_WT.rds")
correlation_pseudotime_target <- readRDS(paste0("results_no_split/correlation_pseudotime_",target,".rds"))
lineages <- read.table("lineage_names.txt")$V1
names(correlation_pseudotime_WT) <- lineages
names(correlation_pseudotime_target) <- lineages
correlation_pseudotime_WT <- correlation_pseudotime_WT[unlist(lapply(correlation_pseudotime_WT, function(x) !(is.null(x))))]
correlation_pseudotime_target <- correlation_pseudotime_target[unlist(lapply(correlation_pseudotime_target, function(x) !(is.null(x))))]

set.seed(44)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE
  
)
options(datatable.fread.datatable=FALSE)
library(SingleCellExperiment)
library(data.table)
library(destiny)
library(scran)
library(scater)
library(batchelor)
```

# Remapping lineage and stage using correlation-based analysis on the specific trajectory only

The original cell-type classification for the chimera data is based on data integration between the entire chimera and the reference atlas. We now refine this by considering only the cells assigned to the limb mesoderm lineage. 

```{r}

atlas_meta <- readRDS("../fromIvan_new_atlas/integrated_meta_celltype_clus.rds")
atlas_sce <- readRDS("../data/big_atlas/big_atlas.rds")
cor_target <- readRDS(paste0("results_no_split/",lineage,"/",lineage,"_cor_cells_",target,".rds"))
cor_WT <- readRDS(paste0("results_no_split/",lineage,"/",lineage,"_cor_cells_WT.rds"))
correlation_pseudotime_lineages_target <- atlas_meta$lineage.clustering[match(names(correlation_pseudotime_target),colnames(atlas_sce))]

chimeraT <- refine_celltype_stage(chimeraTarget,cor_target)
chimeraWT <- refine_celltype_stage(chimeraWT,cor_WT)
```


Differential expression 

```{r }
sce_case <- chimeraTarget
sce_control <- chimeraWT
gene_conv <-  read.table("../data/genes.tsv")
rownames(sce_case) <- gene_conv$V2[match(rownames(sce_case),gene_conv$V1)]
rownames(sce_control) <- gene_conv$V2[match(rownames(sce_control),gene_conv$V1)]
sce_case$perturbed <- sce_case$tomato
sce_control$perturbed <- sce_control$tomato
sce_case <- logNormCounts(sce_case,size.factors=sce_case$sizeFactor)
sce_control <- logNormCounts(sce_control,size.factors=sce_control$sizeFactor)

pseudotime_case <- correlation_pseudotime_target[[lineage]]
pseudotime_control <- correlation_pseudotime_WT[[lineage]]
gene_conv <-  read.table("../data/genes.tsv")
gene_file <- list.files(paste0("results_no_split/",lineage),pattern="genes_sublineage")
temporal_genes <- read.table(paste0("results_no_split/",lineage,"/",gene_file))$V1
temporal_genes <- gene_conv$V2[match(temporal_genes,gene_conv$V1)]

```



```{r}
DE_expression_dynamic_target_lineage <- DE_expression_dynamic(sce_case[temporal_genes,],sce_control[temporal_genes,],pseudotime_case,pseudotime_control)

df_dynamic <- DE_expression_dynamic_target_lineage[,c("dynamic_effect.gene","dynamic_effect_contrasted_with_control.z_value","dynamic_effect_contrasted_with_control.FDR","dynamic_effect.z_value", "dynamic_effect.FDR")]
colnames(df_dynamic) <- c("gene","z_score_contrasted_with_control","FDR_contrast","z_score_case_only","FDR_case_only")
df_dynamic <- df_dynamic[!(is.na(df_dynamic$FDR_contrast)),]
df_dynamic$gene <- sapply(df_dynamic$gene,function(x) strsplit(x,"-")[[1]][1])

dir.create("DE_dynamic_output_files")

write.table(df_dynamic[df_dynamic$FDR_contrast <0.1&df_dynamic$FDR_case_only<0.1,], file=paste0("DE_dynamic_output_files/DE_expression_dynamic_",target,"_",lineage,".csv"),sep=",",row.names=FALSE,col.names=TRUE)

# write.table(df_dynamic, file=paste0("DE_dynamic_output_files/all_DE_expression_dynamic_",target,"_",lineage,".csv"),sep=",",row.names=FALSE,col.names=TRUE)
```

