---
title: "perturbSuite_DE_dynamic summary"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes
---

This notebook summarises DE results from perturbSuite_DE_dynamic for specified chimera (T or Mixl1) and specified lineage. 


# Set-up and loading data and intermediate results

```{r, echo=FALSE, output=FALSE, message=FALSE}
source("core_functions_dynamic.R")
source("core_functions_DA_testing_general.R")
source("chimera_core_functions_big_atlas.R")
library(ggplot2)
library(ggrepel)

options(datatable.fread.datatable=FALSE)
library(SingleCellExperiment)
library(data.table)
library(destiny)
library(scran)
library(scater)
library(batchelor)

knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = c("pdf"),
    dpi=300
)
```



```{r}
files_dynamic_T <- list.files("DE_dynamic_output_files",pattern="DE_expression_dynamic_T")
DE_dynamic_T_all <- data.frame(gene=NULL,z_score_contrasted_with_control=NULL,
                               FDR_contrast=NULL, z_score_case_only=NULL,FDR_case_only=NULL,
                               lineage=NULL)
for (j in 1:length(files_dynamic_T)){
  temp <- read.table(paste0("DE_dynamic_output_files/",files_dynamic_T[j]),sep=",",header=TRUE)
  rownames(temp) <- NULL
  if (nrow(temp) > 0){
    temp$lineage <- strsplit(files_dynamic_T[j],"DE_expression_dynamic_T_")[[1]][2]
    temp$lineage <- strsplit(temp$lineage,".csv")[[1]][1]
    DE_dynamic_T_all <- rbind(DE_dynamic_T_all,temp)
  }
}
DE_dynamic_T_all <- DE_dynamic_T_all[!(is.na(DE_dynamic_T_all$FDR_contrast)),]

#DE_dynamic_T_all
write.table(DE_dynamic_T_all,file="DE_genes_T_dynamic.csv",sep=",",col.names=TRUE,row.names=FALSE)
```

```{r}
files_dynamic_Mixl1 <- list.files("DE_dynamic_output_files",pattern="DE_expression_dynamic_Mixl1")
DE_dynamic_Mixl1_all <- data.frame(gene=NULL,z_score_contrasted_with_control=NULL,
                               FDR_contrast=NULL, z_score_case_only=NULL,FDR_case_only=NULL,
                               lineage=NULL)
for (j in 1:length(files_dynamic_Mixl1)){
  temp <- read.table(paste0("DE_dynamic_output_files/",files_dynamic_Mixl1[j]),sep=",",header=TRUE)
  rownames(temp) <- NULL
  if (nrow(temp) > 0){
    temp$lineage <- strsplit(files_dynamic_Mixl1[j],"DE_expression_dynamic_Mixl1_")[[1]][2]
    temp$lineage <- strsplit(temp$lineage,".csv")[[1]][1]
    DE_dynamic_Mixl1_all <- rbind(DE_dynamic_Mixl1_all,temp)
  }
}

DE_dynamic_Mixl1_all <- DE_dynamic_Mixl1_all[!(is.na(DE_dynamic_Mixl1_all$FDR_contrast)),]
write.table(DE_dynamic_Mixl1_all,file="DE_genes_Mixl1_dynamic.csv",sep=",",col.names=TRUE,row.names=FALSE)
```

We plot the numbers of dynamically DE genes per lineage, first for T and Mixl1 separately and then in a combined plot. 

```{r}
nr_DE_genes_Mixl1 <- table(DE_dynamic_Mixl1_all$lineage)
names(nr_DE_genes_Mixl1) <- sapply(names(nr_DE_genes_Mixl1),function(x) gsub("_"," ",x))
nr_DE_genes_T <- table(DE_dynamic_T_all$lineage)
names(nr_DE_genes_T) <- sapply(names(nr_DE_genes_T),function(x) gsub("_"," ",x))
```



```{r DE_T}
df_DE_T <- data.frame(nr_DE_genes_T=nr_DE_genes_T,
                    lineage=names(nr_DE_genes_T))
ggplot(df_DE_T,aes(x=lineage,y=nr_DE_genes_T)) + geom_bar(stat="identity",fill="orange")+
  theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("T")
```



```{r DE_Mixl1}
df_DE_Mixl1 <- data.frame(nr_DE_genes_Mixl1=nr_DE_genes_Mixl1,
                    lineage=names(nr_DE_genes_Mixl1))
ggplot(df_DE_Mixl1,aes(x=lineage,y=nr_DE_genes_Mixl1)) + geom_bar(stat="identity",fill="purple")+
  theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("Mixl1")
```
Plotting only the lineages with the most dynamically DE genes for the Mixl1 chimeras.


```{r DE_Mixl1_top}
df_DE_Mixl1_top <- df_DE_Mixl1[df_DE_Mixl1$nr_DE_genes_Mixl1.Freq>20,]
df_DE_Mixl1_top <- df_DE_Mixl1_top[order(df_DE_Mixl1_top$nr_DE_genes_Mixl1.Freq,decreasing=TRUE),]
df_DE_Mixl1_top$lineage <- factor(df_DE_Mixl1_top$lineage,levels=df_DE_Mixl1_top$lineage)
ggplot(df_DE_Mixl1_top,aes(x=lineage,y=nr_DE_genes_Mixl1.Freq)) + geom_bar(stat="identity",fill="purple")+
  theme_classic()+ggtitle("Mixl1")+ylab("")+xlab("")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
df_DE <- data.frame(nr_DE_genes=c(nr_DE_genes_T,nr_DE_genes_Mixl1),
                    lineage=c(names(nr_DE_genes_T),names(nr_DE_genes_Mixl1)),
                    target=c(rep("T",length(nr_DE_genes_Mixl1)),
                             rep("Mixl1",length(nr_DE_genes_T))))
ggplot(df_DE,aes(x=lineage,y=nr_DE_genes,fill=target)) + geom_bar(stat="identity")+
  theme_classic()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  scale_fill_manual(values=c("Mixl1"="purple","T"="orange"))
```