---
title: "Differential abundance analysis on cell type and trajectory level"
author: "Magdalena Strauss"
output:
   BiocStyle::html_document:
      code_folding: hide
      number_sections: yes
      toc: yes  
---

This script performs differential abundance analysis for the T and Mixl1 chimera datasets on a cell type and lineage trajectory level. 

```{r}
set.seed(44)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE#,
    #fig.path = "plots_differential_abundance/",
   # dev = c("pdf"),
    #dpi=300
)
library(ggthemes)
suppressPackageStartupMessages(source("core_functions_DA_testing.R"))
suppressPackageStartupMessages(source("chimera_core_functions_big_atlas.R"))
```

Reading in of Waddington-OT scores 

```{r}
library(data.table)
wot_data <- fread("wot_output/trajectory_from_9.25.txt_trajectory.txt",header=TRUE)
atlas_meta  <- readRDS("../fromIvan_new_atlas/integrated_meta_celltype_clus.rds")
columns.keep <- colnames(wot_data)[!(grepl("ExE",colnames(wot_data)))]
wot_data <- as.data.frame(wot_data)
wot_data <- wot_data[intersect(colnames(wot_data),columns.keep)]
```

Loading the chimera data sets. 

```{r}
cells_exclude <- sum(grepl("ExE",atlas_meta$celltype.clustering))#excluding extra-embryonic tissues
chimeraT <- readRDS("../results/chimeraT_2020_extended_mapping_rev.rds")
chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraMixl1 <- readRDS("../results/chimeraMixl1_2020_extended_mapping_rev.rds")
```

# Plotting mapped stages

The following plots are for Supplementary Figure 1. 

```{r plot_mapped_stage_Mixl1,fig.height=2,fig.width=4}
plot_stage(chimeraMixl1,saveFileName="plot_stage_Mixl1_rev.pdf",target="Mixl1",scale="relative") 
```

```{r plot_mapped_stage_T,fig.height=2,fig.width=4}
plot_stage(chimeraT,saveFileName="plot_stage_T_rev.pdf",target="T",scale="relative")
```

```{r plot_mapped_stage_WT,fig.height=2,fig.width=4}
plot_stage(chimeraWT,saveFileName="plot_stage_WT_rev.pdf",target="WT",scale="relative")
```


# Plotting cell types

For Supplementary Figure 1. 

```{r plot_celltype_Mixl1,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraMixl1,saveFileName="plot_celltype_Mixl1_rev.pdf",target="Mixl1") 
```

```{r plot_celltypes_T,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraT,saveFileName="plot_celltype_T_rev.pdf",target="T")
```

```{r plot_celltypes_WT,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraWT,saveFileName="plot_celltype_WT_rev.pdf",target="WT")
```


# Differential abundance per celltype

Cell type-based differential abundance testing (COSICC_group) is performed to determine which cell types are depleted or enriched by T or Mixl1 knockout.

## Mixl1

```{r plot_DA_celltypes_Mixl1,fig.width=8.5,fig.retina=1.05,fig.height=7.5}
da_celltypes_Mixl1 <- da_per_celltype(chimeraMixl1,chimeraWT, "Mixl1")
saveRDS(da_celltypes_Mixl1,file="da_celltypes_Mixl1_rev.rds")
```

## T

```{r,fig.width=8.5,fig.retina=1.05,fig.height=7.5}
da_celltypes_T <- da_per_celltype(chimeraT,chimeraWT, "T")
saveRDS(da_celltypes_T,file="da_celltypes_T_rev.rds")
```

Replotting the figure for T in a different layout.

```{r plot_DA_celltypes_T,fig.width=10,fig.height=6.5}
da_celltypes_T <- readRDS("da_celltypes_T_rev.rds")
alpha <- 0.1
fisher_test_celltypes <- da_celltypes_T
fisher_test_celltypes <- fisher_test_celltypes[order(fisher_test_celltypes$p_values),]
fisher_test_celltypes$sig <- "enriched"
fisher_test_celltypes$sig[ fisher_test_celltypes$odds_ratio < 1] <- "depleted"
fisher_test_celltypes$sig[fisher_test_celltypes$p_values > alpha] <- "not significant"
target <- "T"

p <- ggplot( fisher_test_celltypes , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=celltype)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_celltypes[(apply(cbind(fisher_test_celltypes$odds_ratio,1/fisher_test_celltypes$odds_ratio),1,max)>1.5) | (fisher_test_celltypes$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=celltype), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per celltype enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```


## DA without correcting for experimental bias

To compare, we repeat the celltype-based differential abundance without the bias correction implemented in COSICC_group.

```{r da_without_normalising,fig.width=10,fig.retina=1.05,fig.height=4}
da_celltypes_T_without_normalising <- da_per_celltype_without_normalising(chimeraT,chimeraWT, "T")
da_celltypes_T_without_normalising[order(da_celltypes_T_without_normalising$celltype),]
```


# Differential abundance per celltype and stage

Finding cell types that are enriched or depleted per mapped stage. 

```{r DA_celltype_stage_T,fig.width=9,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_T <- da_per_celltype_per_stage(chimeraT,chimeraWT, "T",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"))

```

Subsetting the plot to cell types found significant differentially abundant overall and related lineages

```{r DA_celltype_stage_T_relevant_celltypes,fig.width=9,fig.retina=1,fig.height=6}
da_per_celltype_per_stage_T <- da_per_celltype_per_stage(chimeraT,chimeraWT, "T",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"),subset_plot = c("Embryo proper endothelium","Dermomyotome" ,"Sclerotome",
  "NMPs","Cranial mesoderm","Venous endothelium","Caudal epiblast","NMPs/Mesoderm-biased","Blood progenitors","Anterior somitic tissues",
  "Notochord","Posterior somitic tissues","Spinal cord progenitors","Embryo proper endothelium","YS endothelium","Erythroid",
  "Allantois endothelium","Lateral plate mesoderm","Mesenchyme","Hindbrain floor plate","YS mesothelium","Intermediate mesoderm",
  "Somitic mesoderm","Presomitic mesoderm","Allantois"))

```

```{r DA_per_celltype_per_stage_Mixl1,fig.width=7,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_Mixl1 <- da_per_celltype_per_stage(chimeraMixl1,chimeraWT, "Mixl1",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"))
```

Subsetting to relevant cell types

```{r DA_per_celltype_per_stage_Mixl1_relevant_celltypes,fig.width=7,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_Mixl1 <- da_per_celltype_per_stage(chimeraMixl1,chimeraWT, "Mixl1",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"),subset_plot=c(da_celltypes_Mixl1$celltype[da_celltypes_Mixl1$sig!="not significant"],
                      unique(da_per_celltype_per_stage_Mixl1$celltype_mapped[da_per_celltype_per_stage_Mixl1$sig]),
                      "Blood progenitors", unique(atlas_meta$celltype.clustering)[grepl("Cardio",atlas_meta$celltype.clustering)]))
```

# Differential abundance for lineage trajectories

Here we apply COSICC_DA_lineage_trajectory to identify lineage trajectories that are enriched or depleted for Mixl1 or T knockout. 

```{r}
wot_data <- wot_data[wot_data$id %in% atlas_meta$cell[!(as.vector(atlas_meta$stage) %in% c("E9.25", "E9.5","mixed_gastrulation"))],]
```


```{r}
celltypes_underscore <- sapply(colnames(wot_data),function(x) gsub("[ ]","_",x))
celltypes_underscore <- sapply(celltypes_underscore,function(x) gsub("/","_",x))
colnames(wot_data) <- celltypes_underscore
rownames(wot_data) <- wot_data$id
wot_data <- wot_data[,-1]
```


For Mixl1 knockout

```{r,fig.width=10,fig.retina=1,fig.height=5}
DT_Mixl1 <- differential_fate_probability(chimeraMixl1,chimeraWT,"Mixl1",wot_data)
saveRDS(DT_Mixl1,file="diff_trajectories_Mixl1_rev.rds")
```

```{r plot_DA_lineages_Mixl1,fig.width=6.5,fig.height=4.5}
da_lineages_Mixl1 <- readRDS("diff_trajectories_Mixl1_rev.rds")
alpha <- 0.1
fisher_test_lineages <- da_lineages_Mixl1
fisher_test_lineages <- fisher_test_lineages[order(fisher_test_lineages$p_values),]
fisher_test_lineages$sig <- "enriched"
fisher_test_lineages$sig[ fisher_test_lineages$odds_ratio < 1] <- "depleted"
fisher_test_lineages$sig[fisher_test_lineages$p_values > alpha] <- "not significant"
target <- "Mixl1"

p <- ggplot( fisher_test_lineages , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=trajectory)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_lineages[(apply(cbind(fisher_test_lineages$odds_ratio,1/fisher_test_lineages$odds_ratio),1,max)>1.5) | (fisher_test_lineages$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=trajectory), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per lineage enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```

For T knockout

```{r,fig.retina=1.05,fig.height=5,fig.width=10}
DT_T <- differential_fate_probability(chimeraT,chimeraWT,"T",wot_data)
saveRDS(DT_T,file="diff_trajectories_T_rev.rds")
```

```{r plot_DA_lineages_T,fig.width=6.5,fig.height=4.5}
da_lineages_T <- readRDS("diff_trajectories_T_rev.rds")
alpha <- 0.1
fisher_test_lineages <- da_lineages_T
fisher_test_lineages <- fisher_test_lineages[order(fisher_test_lineages$p_values),]
fisher_test_lineages$sig <- "enriched"
fisher_test_lineages$sig[ fisher_test_lineages$odds_ratio < 1] <- "depleted"
fisher_test_lineages$sig[fisher_test_lineages$p_values > alpha] <- "not significant"
target <- "T"

p <- ggplot( fisher_test_lineages , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=trajectory)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_lineages[(apply(cbind(fisher_test_lineages$odds_ratio,1/fisher_test_lineages$odds_ratio),1,max)>1.5) | (fisher_test_lineages$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=trajectory), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per lineage enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```


# Comparing differential abundance per trajectory versus differential abuncance per celltype

The plots below illustrate a comparison between abundance of cell type versus abundance of the lineage trajectory leading to the cell type. 

## Mixl1

```{r,fig.height=7,fig.width=10}
da_celltypes_Mixl1 <- readRDS("da_celltypes_Mixl1_rev.rds")
DT_Mixl1 <- readRDS("diff_trajectories_Mixl1_rev.rds")
celltypes_Mixl1_all <- unique(c(gsub("[ ]","_",da_celltypes_Mixl1$celltype),DT_Mixl1$trajectory))
rownames(da_celltypes_Mixl1) <- gsub("[ ]","_",da_celltypes_Mixl1$celltype)
rownames(DT_Mixl1) <- DT_Mixl1$trajectory
odds_ratio_celltype <- rep(NA,length(celltypes_Mixl1_all))
names(odds_ratio_celltype) <- celltypes_Mixl1_all
odds_ratio_celltype[rownames(da_celltypes_Mixl1)] <- da_celltypes_Mixl1$odds_ratio
odds_ratio_trajectory <- rep(NA,length(celltypes_Mixl1_all))
names(odds_ratio_trajectory) <- celltypes_Mixl1_all
odds_ratio_trajectory[rownames(DT_Mixl1)] <- DT_Mixl1$odds_ratio

sig_celltype <- rep("not significant",length(celltypes_Mixl1_all))
names(sig_celltype) <- celltypes_Mixl1_all
sig_celltype[rownames(da_celltypes_Mixl1)] <- da_celltypes_Mixl1$sig
sig_trajectory <- rep("not significant",length(celltypes_Mixl1_all))
names(sig_trajectory) <- celltypes_Mixl1_all
sig_trajectory[rownames(DT_Mixl1)] <- DT_Mixl1$sig

df_Mixl1_celltype_trajectory <- data.frame(celltype_trajectory=celltypes_Mixl1_all,odds_ratio_celltype = odds_ratio_celltype,
odds_ratio_trajectory=odds_ratio_trajectory,
sig_celltype =sig_celltype,sig_trajectory = sig_trajectory)

xx_c <- df_Mixl1_celltype_trajectory$sig_celltype != "not significant"
xx_c[is.na(xx_c)] <- FALSE
xx_t <- df_Mixl1_celltype_trajectory$sig_trajectory != "not significant"
xx_t[is.na(xx_t)] <- FALSE

xx <- rep("not significant",length(xx_c))
xx[xx_t&xx_c] <- "celltype and lineage"
xx[xx_t&(!(xx_c))] <- "lineage"
xx[(!(xx_t))&xx_c] <- "celltype"

yy <- df_Mixl1_celltype_trajectory$odds_ratio_trajectory > 2*df_Mixl1_celltype_trajectory$odds_ratio_celltype
yy[yy] <- "trajectory>2*celltype"
yy[yy==FALSE] <- "neither"
yy[is.na(yy)] <- FALSE
yy[df_Mixl1_celltype_trajectory$odds_ratio_celltype > 2*df_Mixl1_celltype_trajectory$odds_ratio_trajectory] <- "celltype>2*trajectory"
df_Mixl1_celltype_trajectory$any_sig <- xx
df_Mixl1_celltype_trajectory$celltype_trajectory_ratio <- yy

```


```{r Mixl1_celltype_versus_lineage,fig.height=6,fig.width=8}

ggplot(df_Mixl1_celltype_trajectory,aes(x=odds_ratio_celltype,y=odds_ratio_trajectory,color=any_sig)) + geom_point(size=3)+theme_classic(base_size=16)+
  xlab("odds ratio celltype") + ylab("odds ratio trajectory")+
 ggrepel::geom_text_repel(data=df_Mixl1_celltype_trajectory[df_Mixl1_celltype_trajectory$any_sig!="not significant" | df_Mixl1_celltype_trajectory$celltype_trajectory_ratio!="neither",], 
                             aes(x=odds_ratio_celltype, y=odds_ratio_trajectory, label=celltype_trajectory), max.overlaps=Inf, size=4)+
  scale_color_manual(values=c("not significant"="grey","lineage"="purple","celltype"="orange","celltype and lineage" = "blue"))+ 
scale_x_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10)) + scale_y_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10))+
  labs(color="significance", shape="ratio of celltype to\ntrajectory enrichment")+#scale_size_manual(values=c("neither"=1,"trajectory>2*celltype"=2,"celltype>2*trajectory"=3))+ 
      annotate("rect", xmin = Inf, xmax = 1, ymin = Inf, ymax = 1, fill= "blue",alpha=0.1)  + 
      annotate("rect", xmin = -Inf, xmax = 1, ymin = Inf, ymax = 1 , fill= "purple",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = Inf, ymin = 1, ymax = -Inf, fill= "brown",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = -Inf, ymin = -Inf, ymax = 1, fill= "yellow",alpha=0.1) 
```

## T

```{r ,fig.height=7,fig.width=10}
da_celltypes_T <- readRDS("da_celltypes_T_rev.rds")
DT_T <- readRDS("diff_trajectories_T_rev.rds")
celltypes_T_all <- unique(c(gsub("[ ]","_",da_celltypes_T$celltype),DT_T$trajectory))
rownames(da_celltypes_T) <- gsub("[ ]","_",da_celltypes_T$celltype)
rownames(DT_T) <- DT_T$trajectory
odds_ratio_celltype <- rep(NA,length(celltypes_T_all))
names(odds_ratio_celltype) <- celltypes_T_all
odds_ratio_celltype[rownames(da_celltypes_T)] <- da_celltypes_T$odds_ratio
odds_ratio_trajectory <- rep(NA,length(celltypes_T_all))
names(odds_ratio_trajectory) <- celltypes_T_all
odds_ratio_trajectory[rownames(DT_T)] <- DT_T$odds_ratio

sig_celltype <- rep("not significant",length(celltypes_T_all))
names(sig_celltype) <- celltypes_T_all
sig_celltype[rownames(da_celltypes_T)] <- da_celltypes_T$sig
sig_trajectory <- rep("not significant",length(celltypes_T_all))
names(sig_trajectory) <- celltypes_T_all
sig_trajectory[rownames(DT_T)] <- DT_T$sig

df_T_celltype_trajectory <- data.frame(celltype_trajectory=celltypes_T_all,odds_ratio_celltype = odds_ratio_celltype,
odds_ratio_trajectory=odds_ratio_trajectory,
sig_celltype =sig_celltype,sig_trajectory = sig_trajectory)

xx_c <- df_T_celltype_trajectory$sig_celltype != "not significant"
xx_c[is.na(xx_c)] <- FALSE
xx_t <- df_T_celltype_trajectory$sig_trajectory != "not significant"
xx_t[is.na(xx_t)] <- FALSE

xx <- rep("not significant",length(xx_c))
xx[xx_t&xx_c] <- "celltype and lineage"
xx[xx_t&(!(xx_c))] <- "lineage"
xx[(!(xx_t))&xx_c] <- "celltype"

yy <- df_T_celltype_trajectory$odds_ratio_trajectory > 2*df_T_celltype_trajectory$odds_ratio_celltype
yy[yy] <- "trajectory>2*celltype"
yy[yy==FALSE] <- "neither"
yy[is.na(yy)] <- FALSE
yy[df_T_celltype_trajectory$odds_ratio_celltype > 2*df_T_celltype_trajectory$odds_ratio_trajectory] <- "celltype>2*trajectory"
df_T_celltype_trajectory$any_sig <- xx
df_T_celltype_trajectory$celltype_trajectory_ratio <- yy

```


```{r T_celltype_versus_lineage,fig.height=6,fig.width=8}
ggplot(df_T_celltype_trajectory,aes(x=odds_ratio_celltype,y=odds_ratio_trajectory,color=any_sig)) + geom_point(size=3)+theme_classic(base_size=16)+
  xlab("odds ratio celltype") + ylab("odds ratio trajectory")+
 ggrepel::geom_text_repel(data=df_T_celltype_trajectory[df_T_celltype_trajectory$any_sig!="not significant" | df_T_celltype_trajectory$celltype_trajectory_ratio!="neither",], 
                             aes(x=odds_ratio_celltype, y=odds_ratio_trajectory, label=celltype_trajectory), max.overlaps=Inf, size=4)+
  scale_color_manual(values=c("not significant"="grey","lineage"="purple","celltype"="orange","celltype and lineage" = "blue"))+ 
scale_x_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10)) + scale_y_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10))+
  labs(color="significance", shape="ratio of celltype to\ntrajectory enrichment")+#scale_size_manual(values=c("neither"=1,"trajectory>2*celltype"=2,"celltype>2*trajectory"=3))+ 
      annotate("rect", xmin = Inf, xmax = 1, ymin = Inf, ymax = 1, fill= "blue",alpha=0.1)  + 
      annotate("rect", xmin = -Inf, xmax = 1, ymin = Inf, ymax = 1 , fill= "purple",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = Inf, ymin = 1, ymax = -Inf, fill= "brown",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = -Inf, ymin = -Inf, ymax = 1, fill= "yellow",alpha=0.1) 
```


# Per-celltype differential abundance testing - checking replicate consistency

We now run differential abundance analysis for cell types separately for each individual pool and compare the results (Supplementary Figure 6).

## T

```{r ,fig.width=14,fig.retina=1.05,fig.height=6,eval=FALSE}
pools_T <- unique(chimeraT$pool)
da_celltypes_T_replicates <- list()
for (j in 1:length(pools_T)){
  sce_temp <- chimeraT[,chimeraT$pool==pools_T[j]]
  da_celltypes_T_replicates[[j]] <- da_per_celltype(sce_temp,chimeraWT, paste0("T_pool_",toString(pools_T[j])))
}

celltypes <- unlist(lapply(da_celltypes_T_replicates,function(x) x$celltype))
celltypes <- names(table(celltypes))[table(celltypes) == max(table(celltypes))]
sig <- do.call(cbind,lapply(da_celltypes_T_replicates,function(x) x$sig[match(celltypes,x$celltype)]))
sig <- cbind(sig,da_celltypes_T$sig[match(celltypes,da_celltypes_T$celltype)])
rownames(sig) <- celltypes
colnames(sig) <- c(sapply(pools_T,toString),"combined")

odds_ratio <- do.call(cbind,lapply(da_celltypes_T_replicates,function(x) x$odds_ratio[match(celltypes,x$celltype)]))
odds_ratio <- cbind(odds_ratio,da_celltypes_T$odds_ratio[match(celltypes,da_celltypes_T$celltype)])
rownames(odds_ratio) <- celltypes
colnames(odds_ratio) <- c(sapply(pools_T,toString),"combined")
```

```{r,eval=FALSE}
sig
write.table(sig,file="sig_T_rev.csv",col.names=TRUE,row.names=FALSE,sep=",",quote=FALSE)
```

```{r,eval=FALSE}
odds_ratio
write.table(odds_ratio,file="odds_ratio_T_rev.csv",col.names=TRUE,row.names=TRUE,sep=",",quote=FALSE)

```

We compare the odds ratios computed for each pool.

```{r}
odds_ratio_T <- read.table("odds_ratio_T_rev.csv",sep=",",header=TRUE)
```

```{r heatmap_pools_T, fig.height=10}
odds_ratio_T <- read.table("odds_ratio_T_rev.csv",header=TRUE,sep=",")
library(pheatmap)
myBreaks <- c(seq(min(odds_ratio_T), 1, length.out=10), 
              seq(1+max(odds_ratio_T)/20, max(odds_ratio_T), length.out=10))
pheatmap(odds_ratio_T,color=colorRampPalette(c("Darkblue", "white","red"))(20),breaks=myBreaks)
```

We compute the Spearman rank correlations between the odds ratios.  

```{r corr_odds_ratios_pools_T}
pheatmap(cor(odds_ratio_T,method="spearman"),breaks=seq(0,1,by=0.01))
```

We subset to those cell types that are significant for at least one pool or the combined data set. 

```{r}
sig_T <- read.table("sig_T_rev.csv",sep=",",header=TRUE)
odds_ratio_T_sub <- odds_ratio_T[apply(sig_T,1,function(x) any(!(x=="not significant"))),]
```

```{r heatmap_pools_T_sub, fig.height=10}

myBreaks <- c(seq(min(odds_ratio_T_sub), 1, length.out=10), 
              seq(1+max(odds_ratio_T_sub)/20, max(odds_ratio_T_sub), length.out=10))
pheatmap(odds_ratio_T_sub,color=colorRampPalette(c("Darkblue", "white","red"))(20),breaks=myBreaks)
```

We compute the Spearman rank correlations between the odds ratios.  

```{r corr_odds_ratios_pools_T_sub}
pheatmap(cor(odds_ratio_T_sub,method="spearman"),breaks=seq(0,1,by=0.01))
```

## Mixl1

We repeat the above for the Mixl1 chimera data sets. 

```{r ,fig.width=14,fig.retina=1.05,fig.height=6,eval=FALSE}
chimeraMixl1$pool <- floor(chimeraMixl1$sample/2+0.5)
pools_Mixl1 <- unique(chimeraMixl1$pool)
da_celltypes_Mixl1_replicates <- list()
chimeraMixl1$tomato <- as.logical(chimeraMixl1$tomato)
for (j in 1:length(pools_Mixl1)){
  sce_temp<- chimeraMixl1[,chimeraMixl1$pool==pools_Mixl1[j]]
  da_celltypes_Mixl1_replicates[[j]] <- da_per_celltype(sce_temp,chimeraWT, paste0("Mixl1_pool_",toString(pools_Mixl1[j])))
}

celltypes <- unlist(lapply(da_celltypes_Mixl1_replicates,function(x) x$celltype))
celltypes <- names(table(celltypes))[table(celltypes) == max(table(celltypes))]
sig <- do.call(cbind,lapply(da_celltypes_Mixl1_replicates,function(x) x$sig[match(celltypes,x$celltype)]))
sig <- cbind(sig,da_celltypes_Mixl1$sig[match(celltypes,da_celltypes_Mixl1$celltype)])
rownames(sig) <- celltypes
colnames(sig) <- c(sapply(pools_Mixl1,toString),"combined")

odds_ratio <- do.call(cbind,lapply(da_celltypes_Mixl1_replicates,function(x) x$odds_ratio[match(celltypes,x$celltype)]))
odds_ratio <- cbind(odds_ratio,da_celltypes_Mixl1$odds_ratio[match(celltypes,da_celltypes_Mixl1$celltype)])
rownames(odds_ratio) <- celltypes
colnames(odds_ratio) <- c(sapply(pools_Mixl1,toString),"combined")
```

```{r,eval=FALSE}
sig
write.table(sig,file="sig_Mixl1_rev.csv",col.names=TRUE,row.names=FALSE,quote=FALSE,sep=",")
```

```{r,eval=FALSE}
odds_ratio
write.table(odds_ratio,file="odds_ratio_Mixl1_rev.csv",col.names=TRUE,row.names=FALSE,quote=FALSE,sep=",")
```

```{r}
odds_ratio_Mixl1 <- read.table("odds_ratio_Mixl1_rev.csv",sep=",",header=TRUE)
```

```{r heatmap_pools_Mixl1, fig.height=10}
odds_ratio_Mixl1 <- read.table("odds_ratio_Mixl1_rev.csv",header=TRUE,sep=",")
library(pheatmap)
myBreaks <- c(seq(min(odds_ratio_Mixl1), 1, length.out=10), 
              seq(1+max(odds_ratio_Mixl1)/20, max(odds_ratio_Mixl1), length.out=10))
pheatmap(odds_ratio_Mixl1,color=colorRampPalette(c("Darkblue", "white","red"))(20),breaks=myBreaks)
```

We compute the Spearman rank correlations between the odds ratios. 

```{r corr_odds_ratios_pools}
pheatmap(cor(odds_ratio_Mixl1,method="spearman"),breaks=seq(0,1,by=0.01))
```

We subset to those genes that are significant for at least one pool or the combined data set. 

```{r}
sig_Mixl1 <- read.table("sig_Mixl1_rev.csv",sep=",",header=TRUE)
odds_ratio_Mixl1_sub <- odds_ratio_Mixl1[apply(sig_Mixl1,1,function(x) any(!(x=="not significant"))),]
```

```{r heatmap_pools_Mixl1_sub, fig.height=10}

myBreaks <- c(seq(min(odds_ratio_Mixl1_sub), 1, length.out=10), 
              seq(1+max(odds_ratio_Mixl1_sub)/20, max(odds_ratio_Mixl1_sub), length.out=10))
pheatmap(odds_ratio_Mixl1_sub,color=colorRampPalette(c("Darkblue", "white","red"))(20),breaks=myBreaks)
```

```{r corr_odds_ratios_pools_Mixl1_sub}
pheatmap(cor(odds_ratio_Mixl1_sub,method="spearman"),breaks=seq(0,1,by=0.01))
```