---
title: "Calling cells for Mixl1 chimera"
author: "Magdalena Strauss"
output:
  BiocStyle::html_document:
    code_folding: hide
    number_sections: yes
    toc: yes
---
Adapted from https://github.com/MarioniLab/TChimeras2020/tree/master/2_call_cells

```{r seed}
set.seed(42)
```



```{r load, message = FALSE}

library(DropletUtils)
library(ggplot2)
library(Matrix)
library(BiocParallel)
ncores = 6
mcparam = MulticoreParam(workers = ncores)
register(mcparam)
library(reshape2)

setwd("/nfs/research/marioni/magda/chimera")

samples <- c("SIGAA5","SIGAB5","SIGAC5","SIGAD5","SIGAE5","SIGAF5")

mtx_loc = paste0("/nfs/research/marioni/magda/chimera/data_Mixl1/", 
                 samples, "_matrix_unswapped.mtx")
bc_loc =  paste0("/nfs/research/marioni/magda/chimera/data_Mixl1/", 
                 samples, "_barcodes_unswapped.tsv")

#matrices = bplapply(mtx_loc, readMM)
matrices = lapply(mtx_loc, readMM)
bcs = lapply(bc_loc, function(x) read.table(x, header = FALSE, stringsAsFactors = FALSE)[,1])

#correct barcode sample number to match 10X style
for(i in 1:length(bcs)){
  bcs[[i]] = paste0(bcs[[i]], "-", i)
}
```

To calculate the background RNA distribution, we sum the counts in all droplets with fewer than 60 UMIs. Figure \@ref(fig:investigate-edrops) shows that the largest contributions to the background vector derive from small (~40 UMIs) or very small (<5 UMIs) barcodes.

```{r investigate-edrops, fig.cap = "Background vector contributions from barcode libraries of different sizes. Sample 1 is shown, chosen arbitratily."}


mat = matrices[[1]]
targets = mat[, Matrix::colSums(mat)!=0 & Matrix::colSums(mat) < 60]
sums = Matrix::colSums(targets)
reads = as.matrix(table(sums) * as.numeric(names(table(sums))))

# ggplot(data.frame(lib = sums), aes(x = lib)) +
#   geom_histogram(binwidth = 1) +
#   theme_bw() +
#   labs(x = "UMI count")

ggplot(data.frame(lib = reads, n = as.numeric(rownames(reads))), aes(x = n, y= lib)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  labs(x = "Barcode library UMI count", y = "Total contribution to background vector") +
  scale_x_continuous(breaks = seq(0,100, 5))



bg_vec = Matrix::rowSums(targets)


```

The total background vector consists of `r format(sum(reads), big_mark = ",")` UMI counts.


# Calling

Here, we make the calls. The number of called cells per sample is indicated in Table \@ref(tab:save).

We additionally exclude small/low quality (<5000 UMI) barcodes from cell calling.


```{r all-call}

set.seed(42)
#do call
outs = lapply(matrices, emptyDrops, niters = 200000, ignore = 4999, BPPARAM = mcparam, lower = 60)#, retain = Inf)

```


```{r subset-cells}
#identify cells
sigs = lapply(outs, function(x) x$FDR <= 0.01 & !is.na(x$FDR))

#subset the cells
cells = lapply(1:length(matrices), function(i) matrices[[i]][, sigs[[i]]])
barcodes = lapply(1:length(bcs), function(i) bcs[[i]][sigs[[i]]])

#append
counts = do.call(cbind, cells)
barcodes = do.call(c, barcodes)

```

```{r save}
#save
dump_var = writeMM(counts, file = "/nfs/research/marioni/magda/chimera/data/Mixl1/Mixl1_raw_counts.mtx")
write.table(barcodes, file = "/nfs/research/marioni/magda/chimera/data/Mixl1/Mixl1_barcodes.tsv", col.names = FALSE, row.names = FALSE, quote = FALSE)

df_samples = data.frame(sample = 1:length(matrices), value = sapply(cells, ncol))
write.table(df_samples , file = "/nfs/research/marioni/magda/chimera/data/Mixl1/Mixl1_samples.tsv", col.names = FALSE, row.names = FALSE, quote = FALSE)


```

# Session Info
```{r sessinf}
sessionInfo()
```
