---
title: "Dynamic analysis of progression along lineage trajectory"
author: "Magdalena Strauss"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes 
---
  
This notebook computes dynamic shifts caused by T and Mixl1 knockouts for chimeric mouse embryos (COSICC_kinetics). 

```{r}
library(scran)
library(scater)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE#,
    #dev = c("pdf"),
    #dpi=300
)
```

```{r}
chimeraT <- readRDS("../results/chimeraT_2020_extended_mapping_rev.rds")
chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraMixl1 <- readRDS("../results/chimeraMixl1_2020_extended_mapping_rev.rds")
correlation_pseudotime_T <- readRDS("results_no_split/correlation_pseudotime_T.rds")
correlation_pseudotime_WT <- readRDS("results_no_split/correlation_pseudotime_WT.rds")
correlation_pseudotime_Mixl1 <- readRDS("results_no_split/correlation_pseudotime_Mixl1.rds")
Celltypes <- read.table("lineage_names.txt")$V1
names(correlation_pseudotime_T) <- Celltypes
names(correlation_pseudotime_WT) <- Celltypes
names(correlation_pseudotime_Mixl1) <- Celltypes
correlation_pseudotime_T <- correlation_pseudotime_T[unlist(lapply(correlation_pseudotime_T, function(x) !(is.null(x))))]
correlation_pseudotime_WT <- correlation_pseudotime_WT[unlist(lapply(correlation_pseudotime_WT, function(x) !(is.null(x))))]
correlation_pseudotime_Mixl1 <- correlation_pseudotime_Mixl1[unlist(lapply(correlation_pseudotime_Mixl1, function(x) !(is.null(x))))]
```

We perform the tests for lineages for which there is limited depletion of tomato positive cells (the lineage has an odds ratio of at least 0.5). 

```{r}
DT_T <- readRDS("diff_trajectories_T_rev.rds")
DT_Mixl1 <- readRDS("diff_trajectories_Mixl1_rev.rds")
correlation_pseudotime_Mixl1 <- correlation_pseudotime_Mixl1[intersect(names(correlation_pseudotime_Mixl1),DT_Mixl1$trajectory[DT_Mixl1$odds_ratio>0.5])]
correlation_pseudotime_T <-correlation_pseudotime_T[intersect(names(correlation_pseudotime_T),DT_T$trajectory[DT_T$odds_ratio>0.5])]
```

```{r density_Mixl1}
for (j in 1:length(correlation_pseudotime_Mixl1))
  {
  p_Mixl1 <- ggplot(mapping=aes(x= correlation_pseudotime_Mixl1[[j]], 
        fill=chimeraMixl1$tomato[match(names(correlation_pseudotime_Mixl1[[j]]),colnames(chimeraMixl1))])) + 
    geom_density(position = "identity",alpha=0.4) + scale_fill_manual(values=c("darkblue","darkred")) +
    xlab("pseudotime") + ggtitle(paste0("Mixl1 chimera, lineage: ",names(correlation_pseudotime_Mixl1)[j]))+ guides(fill=guide_legend(title="tomato")) + 
                                   theme_classic(base_size=12)+theme(legend.position = "bottom")
 # print(p_Mixl1) #uncomment to plot the distribution of pseudotimes along the lineage trajectory, 
  #coloured by tdTomato (pos. versus negative)
}
```

```{r density_T}
for (j in 1:length(correlation_pseudotime_T))
  {
  p_T <- ggplot(mapping=aes(x= correlation_pseudotime_T[[j]], 
        fill=chimeraT$tomato[match(names(correlation_pseudotime_T[[j]]),colnames(chimeraT))])) + 
    geom_density(position = "identity",alpha=0.4) + scale_fill_manual(values=c("darkblue","darkred")) +
    xlab("pseudotime") + ggtitle(paste0("T chimera, lineage: ",names(correlation_pseudotime_T)[j]))+ guides(fill=guide_legend(title="tomato")) + 
                                   theme_classic(base_size=12)+theme(legend.position = "bottom")
  #print(p_T)  #uncomment to plot the distribution of pseudotimes along the lineage trajectory, 
  #coloured by tdTomato (pos. versus negative
}
```
```{r density_WT}
for (j in 1:length(correlation_pseudotime_WT))
  {
  p_WT <- ggplot(mapping=aes(x= correlation_pseudotime_WT[[j]], 
        fill=chimeraWT$tomato[match(names(correlation_pseudotime_WT[[j]]),colnames(chimeraWT))])) + 
    geom_density(position = "identity",alpha=0.4) + scale_fill_manual(values=c("darkblue","darkred")) +
    xlab("pseudotime") + ggtitle(paste0("WT chimera, lineage: ",names(correlation_pseudotime_WT)[j]))+ guides(fill=guide_legend(title="tomato")) + 
                                   theme_classic(base_size=12)+theme(legend.position = "bottom")
  #print(p_WT)  #uncomment to plot the distribution of pseudotimes along the lineage trajectory, 
  #coloured by tdTomato (pos. versus negative
}
```

We compute the Wilcoxon statistics.

```{r }
df <- data.frame(lower_bound = rep(NA,length(correlation_pseudotime_Mixl1)),upper_bound= rep(NA,length(correlation_pseudotime_Mixl1)),
                 p_value= rep(NA,length(correlation_pseudotime_Mixl1)),estimate= rep(NA,length(correlation_pseudotime_Mixl1)))
for (j in 1:length(correlation_pseudotime_Mixl1)){
  tomato <- chimeraMixl1$tomato[match(names(correlation_pseudotime_Mixl1[[j]]),colnames(chimeraMixl1))]
  wilcox_Mixl1 <- wilcox.test(correlation_pseudotime_Mixl1[[j]][tomato],correlation_pseudotime_Mixl1[[j]][!(tomato)],conf.int = TRUE)
  df$lower_bound[j] <- wilcox_Mixl1$conf.int[1]
  df$upper_bound[j] <- wilcox_Mixl1$conf.int[2]
  df$p_value[j] <- wilcox_Mixl1$p.value
  df$estimate[j] <- wilcox_Mixl1$estimate
}
df$p_value <- p.adjust(df$p_value)
rownames(df) <- names(correlation_pseudotime_Mixl1)
saveRDS(df,file="results_no_split/Wilcoxon_Mixl1.rds")
```


```{r}
df <- data.frame(lower_bound = rep(NA,length(correlation_pseudotime_T)),upper_bound= rep(NA,length(correlation_pseudotime_T)),
                 p_value= rep(NA,length(correlation_pseudotime_T)),estimate= rep(NA,length(correlation_pseudotime_T)))
for (j in 1:length(correlation_pseudotime_T)){
  tomato <- chimeraT$tomato[match(names(correlation_pseudotime_T[[j]]),colnames(chimeraT))]
  wilcox_T <- wilcox.test(correlation_pseudotime_T[[j]][tomato],correlation_pseudotime_T[[j]][!(tomato)],conf.int = TRUE)
  df$lower_bound[j] <- wilcox_T$conf.int[1]
  df$upper_bound[j] <- wilcox_T$conf.int[2]
  df$p_value[j] <- wilcox_T$p.value
  df$estimate[j] <- wilcox_T$estimate
}
df$p_value <- p.adjust(df$p_value)
rownames(df) <- names(correlation_pseudotime_T)
saveRDS(df,file="results_no_split/Wilcoxon_T.rds")
```


```{r}
df <- data.frame(lower_bound = rep(NA,length(correlation_pseudotime_WT)),upper_bound= rep(NA,length(correlation_pseudotime_WT)),
                 p_value= rep(NA,length(correlation_pseudotime_WT)),estimate= rep(NA,length(correlation_pseudotime_WT)))
for (j in 1:length(correlation_pseudotime_WT)){
  tomato <- chimeraWT$tomato[match(names(correlation_pseudotime_WT[[j]]),colnames(chimeraWT))]
  wilcox_WT <- wilcox.test(correlation_pseudotime_WT[[j]][tomato],correlation_pseudotime_WT[[j]][!(tomato)],conf.int = TRUE)
  df$lower_bound[j] <- wilcox_WT$conf.int[1]
  df$upper_bound[j] <- wilcox_WT$conf.int[2]
  df$p_value[j] <- wilcox_WT$p.value
  df$estimate[j] <- wilcox_WT$estimate
}
df$p_value <- p.adjust(df$p_value)
rownames(df) <- names(correlation_pseudotime_WT)
saveRDS(df,file="results_no_split/Wilcoxon_WT.rds")
```

Plotting shift in median between the pseudo-time density of tomato positive versus tomato negative cells. 

```{r Wilcoxon_stats_Mixl1}
Wilcoxon_stats_T <- readRDS("results_no_split/Wilcoxon_T.rds")
Wilcoxon_stats_T$sig <- Wilcoxon_stats_T$p_value < 0.1
DT_T <- readRDS("diff_trajectories_T_rev.rds")

Wilcoxon_stats_WT <- readRDS("results_no_split/Wilcoxon_WT.rds")
Wilcoxon_stats_WT$sig <- Wilcoxon_stats_WT$p_value < 0.1

Wilcoxon_stats_Mixl1 <- readRDS("results_no_split/Wilcoxon_Mixl1.rds")
Wilcoxon_stats_Mixl1$sig <- Wilcoxon_stats_Mixl1$p_value < 0.1
Wilcoxon_stats_Mixl1$location_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_Mixl1),]$estimate
Wilcoxon_stats_Mixl1$lineage <- rownames(Wilcoxon_stats_Mixl1)



Wilcoxon_stats_Mixl1$lower_bound_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_Mixl1),]$lower_bound
Wilcoxon_stats_Mixl1$upper_bound_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_Mixl1),]$upper_bound

ggplot(Wilcoxon_stats_Mixl1,aes(x=location_WT,y=estimate,label=lineage,color=sig)) + geom_point()+geom_hline(yintercept=0) + xlab("location WT")+
  ylab("location Mixl1")+geom_vline(xintercept=0)+ggtitle("") + scale_y_continuous(limits=c(min(Wilcoxon_stats_Mixl1$estimate*1.2),max(Wilcoxon_stats_Mixl1$estimate*1.2)))+
  scale_x_continuous(limits=c(min(Wilcoxon_stats_Mixl1$estimate*1.2),max(Wilcoxon_stats_Mixl1$estimate*1.2))) + theme_bw()+ ggrepel::geom_text_repel(data=Wilcoxon_stats_Mixl1[Wilcoxon_stats_Mixl1$sig,], 
                             aes(x=location_WT, y=estimate, label=lineage), max.overlaps=Inf, size=4) + scale_color_manual(values=c("FALSE"="grey","TRUE"="darkred")) + labs(colour="significant")
```

Plotting significant shifts in the lineage trajectory dynamics 

```{r Wilcoxon_stats_Mixl1_2,fig.height=1.4,fig.width=7}
sig <- rep(FALSE,nrow(Wilcoxon_stats_Mixl1))
xx <-  Wilcoxon_stats_Mixl1$lower_bound_WT > Wilcoxon_stats_Mixl1$upper_bound
yy <- Wilcoxon_stats_Mixl1$upper_bound_WT < Wilcoxon_stats_Mixl1$lower_bound
sig[xx] <- TRUE
sig[yy] <- TRUE
Wilcoxon_stats_Mixl1_sub <- Wilcoxon_stats_Mixl1[Wilcoxon_stats_Mixl1$sig & sig,]
 pp <- ggplot(Wilcoxon_stats_Mixl1_sub, aes(x = lower_bound_WT, y = lineage)) +
    geom_segment(aes(xend = upper_bound_WT, yend = lineage), colour = "black",size=1) +
    geom_point() +
    geom_point(aes(x = upper_bound_WT,show.legend=TRUE)) +
    geom_point(aes(x=lower_bound),color="purple") + 
    geom_point(aes(x=upper_bound),color="purple",show.legend=TRUE) + 
    geom_segment(aes(xend=upper_bound,yend = lineage,x=lower_bound,y=lineage),color="purple",size=1) +
    xlab("conf. interval - Wilcoxon statistic") + theme_bw(base_size=18) + geom_vline(xintercept = 0)+
     annotate("text", x = 0.0005, y = 2.2, label = "Mixl1",color="purple")+ylab("")+
    annotate("text", x = 0.0005, y = 1.7, label = "WT",color="black") +theme_minimal()
  
print(pp)
```

```{r Wilcoxon_stats_T}
Wilcoxon_stats_T$location_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_T),]$estimate
Wilcoxon_stats_T$lineage <- rownames(Wilcoxon_stats_T)

Wilcoxon_stats_T$sig <- Wilcoxon_stats_T$p_value < 0.1
Wilcoxon_stats_T$lower_bound_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_T),]$lower_bound
Wilcoxon_stats_T$upper_bound_WT <- Wilcoxon_stats_WT[rownames(Wilcoxon_stats_T),]$upper_bound

ggplot(Wilcoxon_stats_T,aes(x=location_WT,y=estimate,label=lineage,color=sig)) + geom_point()+geom_hline(yintercept=0) + xlab("location T")+
  ylab("location T")+geom_vline(xintercept=0)+ggtitle("") + scale_y_continuous(limits=c(min(Wilcoxon_stats_T$estimate*1.2),max(Wilcoxon_stats_T$estimate*1.2)))+
  scale_x_continuous(limits=c(min(Wilcoxon_stats_T$estimate*1.2),max(Wilcoxon_stats_T$estimate*1.2))) + theme_bw()+ ggrepel::geom_text_repel(data=Wilcoxon_stats_T[Wilcoxon_stats_T$sig,], 
                             aes(x=location_WT, y=estimate, label=lineage), max.overlaps=Inf, size=4) + scale_color_manual(values=c("FALSE"="grey","TRUE"="darkred")) + labs(colour="significant")
```

```{r Wilcoxon_stats_T_2,fig.height=2,fig.width=7}
sig <- rep(FALSE,nrow(Wilcoxon_stats_T))
xx <-  Wilcoxon_stats_T$lower_bound_WT > Wilcoxon_stats_T$upper_bound
yy <- Wilcoxon_stats_T$upper_bound_WT < Wilcoxon_stats_T$lower_bound
sig[xx] <- TRUE
sig[yy] <- TRUE
Wilcoxon_stats_T_sub <- Wilcoxon_stats_T[Wilcoxon_stats_T$sig & sig,]
 pp <- ggplot(Wilcoxon_stats_T_sub, aes(x = lower_bound_WT, y = lineage)) +
    geom_segment(aes(xend = upper_bound_WT, yend = lineage), colour = "black",size=1) +
    geom_point() +
    geom_point(aes(x = upper_bound_WT,show.legend=TRUE)) +
    geom_point(aes(x=lower_bound),color="purple") + 
    geom_point(aes(x=upper_bound),color="purple",show.legend=TRUE) + 
    geom_segment(aes(xend=upper_bound,yend = lineage,x=lower_bound,y=lineage),color="purple",size=1) +
    xlab("conf. interval - Wilcoxon statistic") + theme_bw(base_size=18) + geom_vline(xintercept = 0)+
    annotate("text", x = 0.005, y = 2.7, label = "T",color="purple")+ylab("")+
    annotate("text", x = 0.005, y = 2.2, label = "WT",color="black") +theme_minimal()
  
print(pp)
```

Summary plots combining enrichment/depletion from the celltype, lineage and trajectory perspective.

```{r ,fig.width=11,fig.height=6}
colours_comp <- c("up in T only"= "#F392E5", "up in Mixl1 only"="#FEED07","up in both"="#FE0707",
"down in T only"="#32CBF6","down in Mixl1 only"="#303B30","down in both"="#1B1FFC",
"up in T, down in Mixl1"="#A24BCE","down in T, up in Mixl1"="#27AC23",
"not significant"="grey")

da_celltypes_T <- readRDS("da_celltypes_T_rev.rds")
da_celltypes_Mixl1 <- readRDS("da_celltypes_Mixl1_rev.rds")
celltypes <- unique(c(da_celltypes_T$celltype,da_celltypes_Mixl1$celltype))
lineages <- unique(c(DT_T$trajectory,DT_Mixl1$trajectory))                    
celltypes <- gsub("[ ]","_",celltypes)
celltypes_not_in_lineage_comparison <- setdiff(celltypes,lineages)
celltypes_not_in_dynamic_comparison_T <- setdiff(celltypes,rownames(Wilcoxon_stats_T))
celltypes_not_in_dynamic_comparison_Mixl1 <- setdiff(celltypes,rownames(Wilcoxon_stats_Mixl1))
lineages_not_in_celltype_comparison <- setdiff(lineages, celltypes)

celltypes <- unique(c(celltypes,lineages))
df_comp <- data.frame(celltype=celltypes)
df_comp$celltype_T <- NA
indDA <- match(gsub("[ ]","_",da_celltypes_T$celltype),celltypes)
df_comp$celltype_T[indDA[!(is.na(indDA))]] <- da_celltypes_T$sig
df_comp$celltype_T[df_comp$celltype%in%lineages_not_in_celltype_comparison] <- "not_included"

df_comp$celltype_Mixl1 <- NA
indDA <- match(gsub("[ ]","_",da_celltypes_Mixl1$celltype),celltypes)
df_comp$celltype_Mixl1[indDA[!(is.na(indDA))]] <- da_celltypes_Mixl1$sig
df_comp$celltype_Mixl1[df_comp$celltype%in%lineages_not_in_celltype_comparison] <- "not_included"

df_comp$lineage_T <- NA
df_comp$lineage_Mixl1 <- NA
indDT <- match(DT_T$trajectory,celltypes)
df_comp$lineage_T[indDT[!(is.na(indDT))]] <- DT_T$sig
df_comp$lineage_T[df_comp$celltype%in%celltypes_not_in_lineage_comparison] <- "not_included"

indDT <- match(DT_Mixl1$trajectory,celltypes)
df_comp$lineage_Mixl1[indDT[!(is.na(indDT))]] <- DT_Mixl1$sig
df_comp$lineage_Mixl1[df_comp$celltype%in%celltypes_not_in_lineage_comparison] <- "not_included"

df_comp$dynamic_trajectory_T <- NA
df_comp$dynamic_trajectory_Mixl1 <- NA

Wilcoxon_stats_T_sub$sig[Wilcoxon_stats_T_sub$upper_bound <  Wilcoxon_stats_T_sub$lower_bound_WT] <- "down"
Wilcoxon_stats_T_sub$sig[Wilcoxon_stats_T_sub$lower_bound >  Wilcoxon_stats_T_sub$upper_bound_WT] <- "up"

indLin <- match(Wilcoxon_stats_T_sub$lineage,df_comp$celltype)
df_comp$dynamic_trajectory_T[indLin[!(is.na(indLin))]] <- Wilcoxon_stats_T_sub$sig
df_comp$dynamic_trajectory_T[df_comp$celltype%in%celltypes_not_in_dynamic_comparison_T] <- "not_included"

Wilcoxon_stats_Mixl1_sub$sig[Wilcoxon_stats_Mixl1_sub$upper_bound <  Wilcoxon_stats_Mixl1_sub$lower_bound_WT] <- "down"
Wilcoxon_stats_Mixl1_sub$sig[Wilcoxon_stats_Mixl1_sub$lower_bound >  Wilcoxon_stats_Mixl1_sub$upper_bound_WT] <- "up"

indLin <- match(Wilcoxon_stats_Mixl1_sub$lineage,df_comp$celltype)
df_comp$dynamic_trajectory_Mixl1[indLin[!(is.na(indLin))]] <- Wilcoxon_stats_Mixl1_sub$sig
df_comp$dynamic_trajectory_Mixl1[df_comp$celltype%in%celltypes_not_in_dynamic_comparison_Mixl1] <- "not_included"

df_comp[is.na(df_comp)] <- "not significant"
df_comp <- df_comp[apply(df_comp,1,function(x) any(!(x%in%c("not significant","not_included")))),]

df_comp_numeric <- matrix(0,nrow=nrow(df_comp),ncol=ncol(df_comp)-1)
colnames(df_comp_numeric) <- colnames(df_comp)[-1]
rownames(df_comp_numeric) <- df_comp[,1]
df_comp_numeric[df_comp[,-1]=="up"] <- 1
df_comp_numeric[df_comp[,-1]=="down"] <- -1
df_comp_numeric[df_comp[,-1]=="enriched"] <- 1
df_comp_numeric[df_comp[,-1]=="depleted"] <- -1
dim(df_comp_numeric) <- c(nrow(df_comp),ncol(df_comp)-1)
colnames(df_comp_numeric) <- colnames(df_comp)[-1]
rownames(df_comp_numeric) <- df_comp[,1]
df_comp_numeric[is.na(df_comp_numeric )] <- 0
df_comp_numeric <- df_comp_numeric[apply(df_comp_numeric,1,function(x) any(x != 0)),]



df_comp$DA_celltype <- "not significant"
df_comp$DA_celltype[df_comp$celltype_T=="depleted" & df_comp$celltype_Mixl1%in%c("not significant","not_included")] <- "down in T only"
df_comp$DA_celltype[df_comp$celltype_Mixl1=="depleted" & df_comp$celltype_T%in%c("not significant","not_included")] <- "down in Mixl1 only"
df_comp$DA_celltype[df_comp$celltype_T=="enriched" & df_comp$celltype_Mixl1%in%c("not significant","not_included")] <- "up in T only"
df_comp$DA_celltype[df_comp$celltype_Mixl1=="enriched" & df_comp$celltype_T%in%c("not significant","not_included")] <- "up in Mixl1 only"
df_comp$DA_celltype[df_comp$celltype_Mixl1=="enriched" & df_comp$celltype_T=="enriched"] <- "up in both"
df_comp$DA_celltype[df_comp$celltype_Mixl1=="depleted" & df_comp$celltype_T=="depleted"] <- "down in both"
df_comp$DA_celltype[df_comp$celltype_Mixl1=="depleted" & df_comp$celltype_T=="enriched"] <- "up in T, down in Mixl1"
df_comp$DA_celltype[df_comp$celltype_T=="depleted" & df_comp$celltype_Mixl1=="enriched"] <- "down in T, up in Mixl1"

df_comp$DA_lineage <- "not significant"
df_comp$DA_lineage[df_comp$lineage_T=="depleted" & df_comp$lineage_Mixl1=="not significant"] <- "down in T only"
df_comp$DA_lineage[df_comp$lineage_Mixl1=="depleted" & df_comp$lineage_T=="not significant"] <- "down in Mixl1 only"
df_comp$DA_lineage[df_comp$lineage_T=="enriched" & df_comp$lineage_Mixl1=="not significant"] <- "up in T only"
df_comp$DA_lineage[df_comp$lineage_Mixl1=="enriched" & df_comp$lineage_T=="not significant"] <- "up in Mixl1 only"
df_comp$DA_lineage[df_comp$lineage_Mixl1=="enriched" & df_comp$lineage_T=="enriched"] <- "up in both"
df_comp$DA_lineage[df_comp$lineage_Mixl1=="depleted" & df_comp$lineage_T=="depleted"] <- "down in both"
df_comp$DA_lineage[df_comp$lineage_Mixl1=="depleted" & df_comp$lineage_T=="enriched"] <- "up in T, down in Mixl1"
df_comp$DA_lineage[df_comp$lineage_T=="depleted" & df_comp$lineage_Mixl1=="enriched"] <- "down in T, up in Mixl1"

df_comp$DA_dynamic_trajectory <- "not significant"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_T=="down" & df_comp$dynamic_trajectory_Mixl1=="not significant"] <- "down in T only"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_Mixl1=="down" & df_comp$dynamic_trajectory_T=="not significant"] <- "down in Mixl1 only"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_T=="up" & df_comp$dynamic_trajectory_Mixl1=="not significant"] <- "up in T only"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_Mixl1=="up" & df_comp$dynamic_trajectory_T=="not significant"] <- "up in Mixl1 only"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_Mixl1=="up" & df_comp$dynamic_trajectory_T=="up"] <- "up in both"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_Mixl1=="down" & df_comp$dynamic_trajectory_T=="down"] <- "down in both"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_Mixl1=="down" & df_comp$dynamic_trajectory_T=="up"] <- "up in T, down in Mixl1"
df_comp$DA_dynamic_trajectory[df_comp$dynamic_trajectory_T=="down" & df_comp$dynamic_trajectory_Mixl1=="up"] <- "down in T, up in Mixl1"


ab <- data.frame(celltype=rep(df_comp$celltype,3),type=factor(c(rep("DA_celltype",nrow(df_comp)),
            rep("DA_lineage",nrow(df_comp)),rep("DA_dynamic_trajectory",nrow(df_comp))),levels=c("DA_celltype","DA_lineage","DA_dynamic_trajectory")),value=c(df_comp$DA_celltype,df_comp$DA_lineage,df_comp$DA_dynamic_trajectory))
xx <- order(ab$value[ab$type=="DA_celltype"])
ab$celltype <- factor(ab$celltype,levels=ab$celltype[ab$type=="DA_celltype"][xx])
ab <- ab[ab$value!="not significant",]
ggplot(ab,mapping=aes(y=celltype,x=type,fill=value)) + geom_raster() + scale_fill_manual(values=colours_comp)


```

The following is a more detailed plot for T chimeras, where we distinguish between cell types that were not significant and those that were not included for a comparison because the cell type does not exist around E8.5 (for cell type level comparisons), the cell type does not exist in large numbers at E9.0 (the end of the trajectory, earlier transitional cell types, for the lineage level comparison), or the lineage is severly depleted in the chimera (for the dynamic analysis).

```{r overview_T, fig.width=5,fig.height=10}
ab <- data.frame(celltype=rep(df_comp$celltype,3),type=factor(c(rep("celltype_T",nrow(df_comp)),
            rep("lineage_T",nrow(df_comp)),rep("dynamic_trajectory_T",nrow(df_comp))),levels=c("celltype_T","lineage_T","dynamic_trajectory_T")),value=c(df_comp$celltype_T,df_comp$lineage_T,df_comp$dynamic_trajectory_T))
ab$celltype <- gsub("_"," ",ab$celltype)
#xx <- order(ab$value[ab$type=="celltype_T"])
xx <- order(ab$celltype)
ab$celltype <- factor(ab$celltype,levels=ab$celltype[ab$type=="celltype_T"][xx])
ab$value[ab$value=="depleted"] <- "down"
ab$value[ab$value=="enriched"] <- "up"
#ab <- ab[ab$value!="not significant",]
ggplot(ab,mapping=aes(y=celltype,x=type,fill=value)) + geom_raster() +
  scale_fill_manual(values=c("not_included" = "white","not significant" = "grey","down"="red","up"="blue"))+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ylab("")
```

Now this plot is repeated for Mixl1. 

```{r overview_Mixl1, fig.width=5,fig.height=10}
ab <- data.frame(celltype=rep(df_comp$celltype,3),type=factor(c(rep("celltype_Mixl1",nrow(df_comp)),
            rep("lineage_Mixl1",nrow(df_comp)),rep("dynamic_trajectory_Mixl1",nrow(df_comp))),levels=c("celltype_Mixl1","lineage_Mixl1","dynamic_trajectory_Mixl1")),value=c(df_comp$celltype_Mixl1,df_comp$lineage_Mixl1,df_comp$dynamic_trajectory_Mixl1))
#xx <- order(ab$value[ab$type=="celltype_Mixl1"])
xx <- order(ab$celltype)
ab$celltype <- gsub("_"," ",ab$celltype)
ab$celltype <- factor(ab$celltype,levels=ab$celltype[ab$type=="celltype_Mixl1"][xx])
ab$value[ab$value=="depleted"] <- "down"
ab$value[ab$value=="enriched"] <- "up"
#ab <- ab[ab$value!="not significant",]
ggplot(ab,mapping=aes(y=celltype,x=type,fill=value)) + geom_raster() +
  scale_fill_manual(values=c("not_included" = "white","not significant" = "grey","down"="red","up"="darkblue"))+ylab("")+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Saving the differential abudance and COSICC_kinetics results as .csv tables. 

```{r}
write.table(da_celltypes_T,"DA_celltype_T.csv",sep=",",col.names=TRUE,row.names=FALSE)
write.table(da_celltypes_Mixl1,"DA_celltype_Mixl1.csv",sep=",",col.names=TRUE,row.names=FALSE)
write.table(DT_T,"DA_lineage_T.csv",sep=",",col.names=TRUE,row.names=FALSE)
write.table(DT_Mixl1,"DA_lineage_Mixl1.csv",sep=",",col.names=TRUE,row.names=FALSE)
```
