---
title: "Differential abundance analysis on cell type and trajectory level"
author: "Magdalena Strauss"
output:
  html_document:
    number_sections: yes
    toc: yes  
    keep_md: yes
---

This script performs differential abundance analysis for the T and Mixl1 chimera datasets on a cell type and trajectory level. 

```{r}
set.seed(44)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    fig.path = "plots_differential_abundance/",
    dev = c("pdf"),
    dpi=300
)
library(ggthemes)
source("core_functions_DA_testing.R")
source("chimera_core_functions_big_atlas.R")
```

Reading in of WOT scores 

```{r}
library(data.table)
wot_data <- fread("wot_output/trajectory_from_9.25.txt_trajectory.txt",header=TRUE)
atlas_meta  <- readRDS("../fromIvan_new_atlas/integrated_meta_celltype_clus.rds")
columns.keep <- colnames(wot_data)[!(grepl("ExE",colnames(wot_data)))]
wot_data <- as.data.frame(wot_data)
wot_data <- wot_data[intersect(colnames(wot_data),columns.keep)]
```

Loading the chimera data sets. 

```{r}
cells_exclude <- sum(grepl("ExE",atlas_meta$celltype.clustering))
chimeraT <- readRDS("../results/chimeraT_2020_extended_mapping_rev.rds")
chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraMixl1 <- readRDS("../results/chimeraMixl1_2020_extended_mapping_rev.rds")
```

# Plotting mapped stages

For supplementary figure

```{r plot_mapped_stage_Mixl1,fig.height=2,fig.width=4}
plot_stage(chimeraMixl1,saveFileName="plot_stage_Mixl1_rev.pdf",target="Mixl1",scale="relative") 
```

```{r plot_mapped_stage_T,fig.height=2,fig.width=4}
plot_stage(chimeraT,saveFileName="plot_stage_T_rev.pdf",target="T",scale="relative")
```

```{r plot_mapped_stage_WT,fig.height=2,fig.width=4}
plot_stage(chimeraWT,saveFileName="plot_stage_WT_rev.pdf",target="WT",scale="relative")
```


# Plotting cell types

For supplementary figure

```{r plot_celltype_Mixl1,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraMixl1,saveFileName="plot_celltype_Mixl1_rev.pdf",target="Mixl1") 
```

```{r plot_celltypes_T,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraT,saveFileName="plot_celltype_T_rev.pdf",target="T")
```

```{r plot_celltypes_WT,fig.height=12,fig.width=8.5}
plot_celltypes(chimeraWT,saveFileName="plot_celltype_WT_rev.pdf",target="WT")
```



# Differential abundance per celltype

Cell type-based differential abundance testing is performed to determine which cell types are depleted or enriched by T or Mixl1 knockout.

## T

```{r,fig.width=8.5,fig.retina=1.05,fig.height=7.5}
da_celltypes_T <- da_per_celltype(chimeraT,chimeraWT, "T")
saveRDS(da_celltypes_T,file="da_celltypes_T_rev.rds")
```

```{r plot_DA_celltypes_T,fig.width=10,fig.height=6.5}
da_celltypes_T <- readRDS("da_celltypes_T_rev.rds")
alpha <- 0.1
fisher_test_celltypes <- da_celltypes_T
fisher_test_celltypes <- fisher_test_celltypes[order(fisher_test_celltypes$p_values),]
fisher_test_celltypes$sig <- "enriched"
fisher_test_celltypes$sig[ fisher_test_celltypes$odds_ratio < 1] <- "depleted"
fisher_test_celltypes$sig[fisher_test_celltypes$p_values > alpha] <- "not significant"
target <- "T"

p <- ggplot( fisher_test_celltypes , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=celltype)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_celltypes[(apply(cbind(fisher_test_celltypes$odds_ratio,1/fisher_test_celltypes$odds_ratio),1,max)>1.5) | (fisher_test_celltypes$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=celltype), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per celltype enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```


## Mixl1

```{r plot_DA_celltypes_Mixl1,fig.width=8.5,fig.retina=1.05,fig.height=7.5}
da_celltypes_Mixl1 <- da_per_celltype(chimeraMixl1,chimeraWT, "Mixl1")
saveRDS(da_celltypes_Mixl1,file="da_celltypes_Mixl1_rev.rds")
```

To compare, we repeat the celltype-based differential abundance without normalising (supplementary figure)

```{r da_without_normalising,fig.width=10,fig.retina=1.05,fig.height=4}
da_celltypes_T_without_normalising <- da_per_celltype_without_normalising(chimeraT,chimeraWT, "T")
da_celltypes_T_without_normalising[order(da_celltypes_T_without_normalising$celltype),]
```

# Differential abundance per celltype using the 2020 celltype annotations 

```{r}
chimeraT_2020 <- readRDS("chimeraT_original_mapping.rds")
chimeraWT_2020 <- readRDS("chimeraWT_original_mapping.rds")
chimeraT_2020 <- chimeraT_2020[,chimeraT_2020$stage=="E8.5"]
chimeraWT_2020 <- chimeraWT_2020[,chimeraWT_2020$stage=="E8.5"]
yy <- complete.cases(reducedDims(chimeraWT_2020)$pca.corrected.E8.5)
chimeraWT_2020 <- chimeraWT_2020[, yy]
yy <- complete.cases(reducedDims(chimeraT_2020)$pca.corrected.E8.5)
chimeraT_2020 <- chimeraT_2020[,yy]
```

```{r da_per_celltype_2020,fig.width=9,fig.retina=1.05,fig.height=3}
da_celltypes_T_2020 <- da_per_celltype(chimeraT_2020,chimeraWT_2020, "T_2020")
da_celltypes_T_2020[order(da_celltypes_T_2020$celltype),]
```



# Differential abundance per celltype and stage

Finding cell types that are enriched or depleted per mapped stage. 

```{r DA_celltype_stage_T,fig.width=9,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_T <- da_per_celltype_per_stage(chimeraT,chimeraWT, "T",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"))

```

Subsetting the plot to cell types found significant overall and related lineages

```{r DA_celltype_stage_T_relevant_celltypes,fig.width=9,fig.retina=1,fig.height=6}
da_per_celltype_per_stage_T <- da_per_celltype_per_stage(chimeraT,chimeraWT, "T",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"),subset_plot = c("Embryo proper endothelium","Dermomyotome" ,"Sclerotome",
  "NMPs","Cranial mesoderm","Venous endothelium","Caudal epiblast","NMPs/Mesoderm-biased","Blood progenitors","Anterior somitic tissues",
  "Notochord","Posterior somitic tissues","Spinal cord progenitors","Embryo proper endothelium","YS endothelium","Erythroid",
  "Allantois endothelium","Lateral plate mesoderm","Mesenchyme","Hindbrain floor plate","YS mesothelium","Intermediate mesoderm",
  "Somitic mesoderm","Presomitic mesoderm","Allantois"))

```

```{r DA_per_celltype_per_stage_Mixl1,fig.width=7,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_Mixl1 <- da_per_celltype_per_stage(chimeraMixl1,chimeraWT, "Mixl1",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"))
```

Subsetting to relevant cell types

```{r DA_per_celltype_per_stage_Mixl1_relevant_celltypes,fig.width=7,fig.retina=1,fig.height=9}
da_per_celltype_per_stage_Mixl1 <- da_per_celltype_per_stage(chimeraMixl1,chimeraWT, "Mixl1",stages=c("E7.75","E8.0",
                            "E8.25","E8.5","E8.75"),subset_plot=c(da_celltypes_Mixl1$celltype[da_celltypes_Mixl1$sig!="not significant"],
                      unique(da_per_celltype_per_stage_Mixl1$celltype_mapped[da_per_celltype_per_stage_Mixl1$sig]),
                      "Blood progenitors", unique(atlas_meta$celltype.clustering)[grepl("Cardio",atlas_meta$celltype.clustering)]))
```

# Differential abundance for trajectories


```{r}
wot_data <- wot_data[wot_data$id %in% atlas_meta$cell[!(as.vector(atlas_meta$stage) %in% c("E9.25", "E9.5","mixed_gastrulation"))],]
```


```{r}
celltypes_underscore <- sapply(colnames(wot_data),function(x) gsub("[ ]","_",x))
celltypes_underscore <- sapply(celltypes_underscore,function(x) gsub("/","_",x))
colnames(wot_data) <- celltypes_underscore
rownames(wot_data) <- wot_data$id
wot_data <- wot_data[,-1]
```


For Mixl1

```{r,fig.width=10,fig.retina=1,fig.height=5}
DT_Mixl1 <- differential_fate_probability(chimeraMixl1,chimeraWT,"Mixl1",wot_data)
saveRDS(DT_Mixl1,file="diff_trajectories_Mixl1_rev.rds")
```

```{r plot_DA_lineages_Mixl1,fig.width=6.5,fig.height=4.5}
da_lineages_Mixl1 <- readRDS("diff_trajectories_Mixl1_rev.rds")
alpha <- 0.1
fisher_test_lineages <- da_lineages_Mixl1
fisher_test_lineages <- fisher_test_lineages[order(fisher_test_lineages$p_values),]
fisher_test_lineages$sig <- "enriched"
fisher_test_lineages$sig[ fisher_test_lineages$odds_ratio < 1] <- "depleted"
fisher_test_lineages$sig[fisher_test_lineages$p_values > alpha] <- "not significant"
target <- "Mixl1"

p <- ggplot( fisher_test_lineages , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=trajectory)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_lineages[(apply(cbind(fisher_test_lineages$odds_ratio,1/fisher_test_lineages$odds_ratio),1,max)>1.5) | (fisher_test_lineages$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=trajectory), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per lineage enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```

For T

```{r,fig.retina=1.05,fig.height=5,fig.width=10}
DT_T <- differential_fate_probability(chimeraT,chimeraWT,"T",wot_data)
saveRDS(DT_T,file="diff_trajectories_T_rev.rds")
```

```{r plot_DA_lineages_T,fig.width=6.5,fig.height=4.5}
da_lineages_T <- readRDS("diff_trajectories_T_rev.rds")
alpha <- 0.1
fisher_test_lineages <- da_lineages_T
fisher_test_lineages <- fisher_test_lineages[order(fisher_test_lineages$p_values),]
fisher_test_lineages$sig <- "enriched"
fisher_test_lineages$sig[ fisher_test_lineages$odds_ratio < 1] <- "depleted"
fisher_test_lineages$sig[fisher_test_lineages$p_values > alpha] <- "not significant"
target <- "T"

p <- ggplot( fisher_test_lineages , aes(x=log10(odds_ratio), y=-log10(p_values+1e-100),text=trajectory)) +
  labs(x=expression(paste("log"[10]," of odds ratio")),y=expression(paste("-log"[10]," of FDR adjusted p-values"))) +
  geom_point(aes(color=sig),size=3) +
  scale_color_manual(values=c("not significant"="grey","enriched"="darkblue","depleted" = "darkred"),name="") +
  ggrepel::geom_text_repel(data=fisher_test_lineages[(apply(cbind(fisher_test_lineages$odds_ratio,1/fisher_test_lineages$odds_ratio),1,max)>1.5) | (fisher_test_lineages$p_values < 0.1),], 
                           aes(x=log10(odds_ratio), y=-log10(p_values+1e-100), label=trajectory), max.overlaps=Inf, size=4) +
  theme_classic(base_size=14) + scale_y_continuous(trans=pseudo_log_trans(sigma=1,base=10))+
  theme(
    axis.text = element_text(size=rel(0.75), color='black'),
    axis.title = element_text(size=rel(1.0), color='black')) + ggtitle(paste0(target,"- per lineage enrichment"))+
  annotate("rect", xmin = Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "blue",alpha=0.1) +
  annotate("rect", xmin = -Inf, xmax = 0, ymin = Inf, ymax = -Inf, fill= "red",alpha=0.1) 
print(p)
```


# Comparing differential abundance per trajectory versus differential abuncance per celltype

## Mixl1

```{r,fig.height=7,fig.width=10}
da_celltypes_Mixl1 <- readRDS("da_celltypes_Mixl1_rev.rds")
DT_Mixl1 <- readRDS("diff_trajectories_Mixl1_rev.rds")
celltypes_Mixl1_all <- unique(c(gsub("[ ]","_",da_celltypes_Mixl1$celltype),DT_Mixl1$trajectory))
rownames(da_celltypes_Mixl1) <- gsub("[ ]","_",da_celltypes_Mixl1$celltype)
rownames(DT_Mixl1) <- DT_Mixl1$trajectory
odds_ratio_celltype <- rep(NA,length(celltypes_Mixl1_all))
names(odds_ratio_celltype) <- celltypes_Mixl1_all
odds_ratio_celltype[rownames(da_celltypes_Mixl1)] <- da_celltypes_Mixl1$odds_ratio
odds_ratio_trajectory <- rep(NA,length(celltypes_Mixl1_all))
names(odds_ratio_trajectory) <- celltypes_Mixl1_all
odds_ratio_trajectory[rownames(DT_Mixl1)] <- DT_Mixl1$odds_ratio

sig_celltype <- rep("not significant",length(celltypes_Mixl1_all))
names(sig_celltype) <- celltypes_Mixl1_all
sig_celltype[rownames(da_celltypes_Mixl1)] <- da_celltypes_Mixl1$sig
sig_trajectory <- rep("not significant",length(celltypes_Mixl1_all))
names(sig_trajectory) <- celltypes_Mixl1_all
sig_trajectory[rownames(DT_Mixl1)] <- DT_Mixl1$sig

df_Mixl1_celltype_trajectory <- data.frame(celltype_trajectory=celltypes_Mixl1_all,odds_ratio_celltype = odds_ratio_celltype,
odds_ratio_trajectory=odds_ratio_trajectory,
sig_celltype =sig_celltype,sig_trajectory = sig_trajectory)

xx_c <- df_Mixl1_celltype_trajectory$sig_celltype != "not significant"
xx_c[is.na(xx_c)] <- FALSE
xx_t <- df_Mixl1_celltype_trajectory$sig_trajectory != "not significant"
xx_t[is.na(xx_t)] <- FALSE

xx <- rep("not significant",length(xx_c))
xx[xx_t&xx_c] <- "celltype and lineage"
xx[xx_t&(!(xx_c))] <- "lineage"
xx[(!(xx_t))&xx_c] <- "celltype"

yy <- df_Mixl1_celltype_trajectory$odds_ratio_trajectory > 2*df_Mixl1_celltype_trajectory$odds_ratio_celltype
yy[yy] <- "trajectory>2*celltype"
yy[yy==FALSE] <- "neither"
yy[is.na(yy)] <- FALSE
yy[df_Mixl1_celltype_trajectory$odds_ratio_celltype > 2*df_Mixl1_celltype_trajectory$odds_ratio_trajectory] <- "celltype>2*trajectory"
df_Mixl1_celltype_trajectory$any_sig <- xx
df_Mixl1_celltype_trajectory$celltype_trajectory_ratio <- yy

```


```{r Mixl1_celltype_versus_lineage,fig.height=6,fig.width=8}

ggplot(df_Mixl1_celltype_trajectory,aes(x=odds_ratio_celltype,y=odds_ratio_trajectory,color=any_sig)) + geom_point(size=3)+theme_classic(base_size=16)+
  xlab("odds ratio celltype") + ylab("odds ratio trajectory")+
 ggrepel::geom_text_repel(data=df_Mixl1_celltype_trajectory[df_Mixl1_celltype_trajectory$any_sig!="not significant" | df_Mixl1_celltype_trajectory$celltype_trajectory_ratio!="neither",], 
                             aes(x=odds_ratio_celltype, y=odds_ratio_trajectory, label=celltype_trajectory), max.overlaps=Inf, size=4)+
  scale_color_manual(values=c("not significant"="grey","lineage"="purple","celltype"="orange","celltype and lineage" = "blue"))+ 
scale_x_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10)) + scale_y_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10))+
  labs(color="significance", shape="ratio of celltype to\ntrajectory enrichment")+#scale_size_manual(values=c("neither"=1,"trajectory>2*celltype"=2,"celltype>2*trajectory"=3))+ 
      annotate("rect", xmin = Inf, xmax = 1, ymin = Inf, ymax = 1, fill= "blue",alpha=0.1)  + 
      annotate("rect", xmin = -Inf, xmax = 1, ymin = Inf, ymax = 1 , fill= "purple",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = Inf, ymin = 1, ymax = -Inf, fill= "brown",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = -Inf, ymin = -Inf, ymax = 1, fill= "yellow",alpha=0.1) 
```

## T

```{r ,fig.height=7,fig.width=10}
da_celltypes_T <- readRDS("da_celltypes_T_rev.rds")
DT_T <- readRDS("diff_trajectories_T_rev.rds")
celltypes_T_all <- unique(c(gsub("[ ]","_",da_celltypes_T$celltype),DT_T$trajectory))
rownames(da_celltypes_T) <- gsub("[ ]","_",da_celltypes_T$celltype)
rownames(DT_T) <- DT_T$trajectory
odds_ratio_celltype <- rep(NA,length(celltypes_T_all))
names(odds_ratio_celltype) <- celltypes_T_all
odds_ratio_celltype[rownames(da_celltypes_T)] <- da_celltypes_T$odds_ratio
odds_ratio_trajectory <- rep(NA,length(celltypes_T_all))
names(odds_ratio_trajectory) <- celltypes_T_all
odds_ratio_trajectory[rownames(DT_T)] <- DT_T$odds_ratio

sig_celltype <- rep("not significant",length(celltypes_T_all))
names(sig_celltype) <- celltypes_T_all
sig_celltype[rownames(da_celltypes_T)] <- da_celltypes_T$sig
sig_trajectory <- rep("not significant",length(celltypes_T_all))
names(sig_trajectory) <- celltypes_T_all
sig_trajectory[rownames(DT_T)] <- DT_T$sig

df_T_celltype_trajectory <- data.frame(celltype_trajectory=celltypes_T_all,odds_ratio_celltype = odds_ratio_celltype,
odds_ratio_trajectory=odds_ratio_trajectory,
sig_celltype =sig_celltype,sig_trajectory = sig_trajectory)

xx_c <- df_T_celltype_trajectory$sig_celltype != "not significant"
xx_c[is.na(xx_c)] <- FALSE
xx_t <- df_T_celltype_trajectory$sig_trajectory != "not significant"
xx_t[is.na(xx_t)] <- FALSE

xx <- rep("not significant",length(xx_c))
xx[xx_t&xx_c] <- "celltype and lineage"
xx[xx_t&(!(xx_c))] <- "lineage"
xx[(!(xx_t))&xx_c] <- "celltype"

yy <- df_T_celltype_trajectory$odds_ratio_trajectory > 2*df_T_celltype_trajectory$odds_ratio_celltype
yy[yy] <- "trajectory>2*celltype"
yy[yy==FALSE] <- "neither"
yy[is.na(yy)] <- FALSE
yy[df_T_celltype_trajectory$odds_ratio_celltype > 2*df_T_celltype_trajectory$odds_ratio_trajectory] <- "celltype>2*trajectory"
df_T_celltype_trajectory$any_sig <- xx
df_T_celltype_trajectory$celltype_trajectory_ratio <- yy

```


```{r T_celltype_versus_lineage,fig.height=6,fig.width=8}
ggplot(df_T_celltype_trajectory,aes(x=odds_ratio_celltype,y=odds_ratio_trajectory,color=any_sig)) + geom_point(size=3)+theme_classic(base_size=16)+
  xlab("odds ratio celltype") + ylab("odds ratio trajectory")+
 ggrepel::geom_text_repel(data=df_T_celltype_trajectory[df_T_celltype_trajectory$any_sig!="not significant" | df_T_celltype_trajectory$celltype_trajectory_ratio!="neither",], 
                             aes(x=odds_ratio_celltype, y=odds_ratio_trajectory, label=celltype_trajectory), max.overlaps=Inf, size=4)+
  scale_color_manual(values=c("not significant"="grey","lineage"="purple","celltype"="orange","celltype and lineage" = "blue"))+ 
scale_x_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10)) + scale_y_continuous(expand = c(0, 0),limits=c(0,3),trans=scales::pseudo_log_trans(sigma=0.1,base = 10))+
  labs(color="significance", shape="ratio of celltype to\ntrajectory enrichment")+#scale_size_manual(values=c("neither"=1,"trajectory>2*celltype"=2,"celltype>2*trajectory"=3))+ 
      annotate("rect", xmin = Inf, xmax = 1, ymin = Inf, ymax = 1, fill= "blue",alpha=0.1)  + 
      annotate("rect", xmin = -Inf, xmax = 1, ymin = Inf, ymax = 1 , fill= "purple",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = Inf, ymin = 1, ymax = -Inf, fill= "brown",alpha=0.1) + 
      annotate("rect", xmin = 1, xmax = -Inf, ymin = -Inf, ymax = 1, fill= "yellow",alpha=0.1) 
```


