---
title: "Additional plots for Figure 1 and Supplementary Figure 1"
author: "Magdalena Strauss"
output:
  html_document:
    keep_md: true
---

This script is the R code used to create figure panels for Figure 1 and Supplementary Figure 1 not created by other scripts. 

```{r}
set.seed(44)
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE,
    message = FALSE,
    dev = c("pdf","png"),
    dpi=300
)
library(ggthemes)
library(ggpubr)
.libPaths("/hps/software/users/marioni/magda/miniconda3/envs/R4.1/lib/R/library")
source("chimera_core_functions_big_atlas.R")
library(patchwork)
```

We plot percentage donut plots for the figure panel illustrating the sampling bias correction. 

```{r}
plot_donut_percentage <- function(percentage){
  data <- data.frame(
  category=c("perturbed", "unperturbed"),
  count=c(percentage,100-percentage))
  data$fraction = data$count/sum(data$count)
  data$theta_max = cumsum(data$fraction)
  data$theta_min = c(0, head(data$theta_max, n=-1))
  data$position_label <- (data$theta_min + data$theta_max)*0.5
  data$label <- paste0(data$count,"%")

  ggplot(data, aes(ymax=theta_max, ymin=theta_min, xmax=2, xmin=1, fill=category)) +
    geom_rect() +
    geom_label( x=1.5, aes(y=position_label, label=label), size=10,color="white") +
    scale_fill_manual(values=c("darkred","darkblue")) +
    coord_polar(theta="y") +
    xlim(c(0.35, 2)) +
    theme_void() +
    theme(legend.position = "none")
 
}
```
```{r, donut_50}
plot_donut_percentage(50)
```
```{r, donut_35}
plot_donut_percentage(35)
```

```{r, donut_0}
plot_donut_percentage(0)
```

```{r, donut_59}
plot_donut_percentage(59)
```

```{r}
atlas_sce <- readRDS("../data/big_atlas/big_atlas.rds")
atlas_meta  <- readRDS("../fromIvan_new_atlas/integrated_meta_celltype_clus.rds")
atlas_meta <- atlas_meta[match(colnames(atlas_sce),atlas_meta$cell),]
atlas_sce$celltype.clustering <- atlas_meta$celltype.clustering
atlas_sce$stage <- atlas_meta$stage
atlas_sce$mapping <- "atlas"
atlas_sce$nr_cells <- 1
chimeraT <- readRDS("../results/chimeraT_2020_extended_mapping_rev.rds")
chimeraWT <- readRDS("../results/chimeraWT_2020_extended_mapping_rev.rds")
chimeraMixl1 <- readRDS("../results/chimeraMixl1_2020_extended_mapping_rev.rds")
chimera_Mixl1_tomato_pos <- chimeraT[,chimeraT$tomato]
chimeraT_tomato_neg <- chimeraT[,!(chimeraT$tomato)]
chimera_T_tomato_pos <- chimeraT[,chimeraT$tomato]
chimera_WT_tomato_pos <- chimeraWT[,chimeraWT$tomato]
chimeraWT_tomato_neg <- chimeraWT[,!(chimeraWT$tomato)]
chimera_Mixl1_tomato_pos <- chimeraMixl1[,chimeraMixl1$tomato]
chimeraMixl1_tomato_neg <- chimeraMixl1[,!(chimeraMixl1$tomato)]
atlas_sce$T_expression <- logcounts(atlas_sce)["ENSMUSG00000062327",]
atlas_sce$Mixl1_expression <- logcounts(atlas_sce)["ENSMUSG00000026497",]
```

We plot UMAPs for the extended mouse gastrulation atlas labelled by cell type.

```{r atlas_celltypes,dev='png'}
p_atlas_celltype <- plotReducedDim(atlas_sce[,sample(1:ncol(atlas_sce))],colour_by = "celltype.clustering",dimred = "umap",point_size=1) + scale_color_manual(values=celltype_colours_final)+theme(legend.position = "bottom",
            legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank()) + guides(color=guide_legend(ncol=4))+theme(legend.position = "None") 
print(p_atlas_celltype)
```

```{r atlas_celltypes_legend,dev='pdf'}
legend_atlas_celltype <- get_legend(p_atlas_celltype)
as_ggplot(legend_atlas_celltype)
```

```{r atlas_celltypes_with_legend,dev=c('pdf','svg')}
p_atlas_celltype <- plotReducedDim(atlas_sce[,sample(1:ncol(atlas_sce))],colour_by = "celltype.clustering",dimred = "umap",point_size=1) + scale_color_manual(values=celltype_colours_final)+theme(legend.position = "bottom",
            legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank()) + guides(color=guide_legend(ncol=4))+theme(legend.position = "Bottom") 
print(p_atlas_celltype)
```

We repeat the UMAP plots coloured by stage. 

```{r atlas_stage, dev='png'}
p_atlas_stage <- plotReducedDim(atlas_sce[,sample(1:ncol(atlas_sce))],colour_by = "stage",dimred = "umap",point_size=1) + theme(legend.position = "bottom",
                                                                              legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank()) + scale_color_viridis_d()
print(p_atlas_stage)
# 
# print(p_atlas_stage)
# pdf(paste0(fig_path,"p_atlas_stage.pdf"))
# print(p_atlas_stage)
# png(paste0(fig_path,"p_atlas_stage.png"))
```

```{r atlas_stage_legend,dev='pdf'}
legend_atlas_stage <- get_legend(p_atlas_stage)
as_ggplot(legend_atlas_stage)
```

We plot nearest neighbours of chimera cells in atlas, first for WT.

```{r mapping_WT_chimeras, dev='png'}
cells <- chimeraWT$closest.cell
atlas_sub <- atlas_sce[,colnames(atlas_sce)%in%cells]
nr_WT_positive <- table(chimera_WT_tomato_pos$closest.cell[chimera_WT_tomato_pos$closest.cell%in%colnames(atlas_sub)])
nr_WT_negative <- table(chimeraWT_tomato_neg$closest.cell[chimeraWT_tomato_neg$closest.cell%in%colnames(atlas_sub)])
atlas_WT_tomato_pos <- atlas_sub[,colnames(atlas_sub)%in%chimera_WT_tomato_pos$closest.cell]
atlas_WT_tomato_pos$nr_cells <- nr_WT_positive[colnames(atlas_WT_tomato_pos)]
atlas_WT_tomato_neg <- atlas_sub[,colnames(atlas_sub)%in%chimeraWT_tomato_neg$closest.cell]
atlas_WT_tomato_neg$nr_cells <- nr_WT_negative[colnames(atlas_WT_tomato_neg)]
sce_temp <- cbind(atlas_WT_tomato_pos,atlas_WT_tomato_neg)
sce_temp$mapping <- c( rep("WT tomato pos",ncol(atlas_WT_tomato_pos)),
                      rep("WT tomato neg",ncol(atlas_WT_tomato_neg)))
atlas_sub$mapping <- "atlas"
atlas_sub$nr_cells <- 1
sce_temp <- sce_temp[,sample(1:ncol(sce_temp))]
sce_temp_3 <- cbind(atlas_sce,sce_temp)
  
  
p3 <- ggplot(mapping=aes(x=reducedDims(sce_temp_3)$umap[,1],
                   y=reducedDims(sce_temp_3)$umap[,2],
                   color=sce_temp_3$mapping,size=sce_temp_3$nr_cells)) + geom_jitter() + theme_classic(base_size=10)+ggtitle("WT")+
  scale_color_manual(values=c( "WT tomato neg"="darkblue","WT tomato pos"="red","atlas"="lightgrey"))+xlab("umap 1")+ylab("umap  2") + theme(legend.position = "none", legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),legend.text=element_text(size=12))

print(p3)
```

```{r mapping_WT_chimeras_legend,dev='svg',dev='pdf'}
legend_mapping_WT_chimeras <- get_legend(p3)
as_ggplot(legend_mapping_WT_chimeras)
```

Now for T

```{r mapping_T_chimeras, dev='png'}
cells <- chimeraT$closest.cell
atlas_sub <- atlas_sce[,colnames(atlas_sce)%in%cells]
nr_T_positive <- table(chimera_T_tomato_pos$closest.cell[chimera_T_tomato_pos$closest.cell%in%colnames(atlas_sub)])
nr_T_negative <- table(chimeraT_tomato_neg$closest.cell[chimeraT_tomato_neg$closest.cell%in%colnames(atlas_sub)])
atlas_T_tomato_pos <- atlas_sub[,colnames(atlas_sub)%in%chimera_T_tomato_pos$closest.cell]
atlas_T_tomato_pos$nr_cells <- nr_T_positive[colnames(atlas_T_tomato_pos)]
atlas_T_tomato_neg <- atlas_sub[,colnames(atlas_sub)%in%chimeraT_tomato_neg$closest.cell]
atlas_T_tomato_neg$nr_cells <- nr_T_negative[colnames(atlas_T_tomato_neg)]
sce_temp <- cbind(atlas_T_tomato_pos,atlas_T_tomato_neg)
sce_temp$mapping <- c( rep("T tomato pos",ncol(atlas_T_tomato_pos)),
                      rep("T tomato neg",ncol(atlas_T_tomato_neg)))
atlas_sub$mapping <- "atlas"
atlas_sub$nr_cells <- 1
sce_temp <- sce_temp[,sample(1:ncol(sce_temp))]
sce_temp_3 <- cbind(atlas_sce,sce_temp)
  
  
p3 <- ggplot(mapping=aes(x=reducedDims(sce_temp_3)$umap[,1],
                   y=reducedDims(sce_temp_3)$umap[,2],
                   color=sce_temp_3$mapping,size=sce_temp_3$nr_cells)) + geom_jitter() + theme_classic(base_size=10)+ggtitle("T")+
  scale_color_manual(values=c( "T tomato neg"="darkblue","T tomato pos"="red","atlas"="lightgrey"))+xlab("umap 1")+ylab("umap 2") + theme(legend.position = "none", legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),legend.text=element_text(size=12))

print(p3)
```

```{r mapping_T_chimeras_legend,dev='svg',dev='pdf'}
legend_mapping_T_chimeras <- get_legend(p3)
as_ggplot(legend_mapping_T_chimeras)
```

And finally for Mixl1 chimeras

```{r mapping_Mixl1_chimeras, dev='png'}
cells <- chimeraMixl1$closest.cell
atlas_sub <- atlas_sce[,colnames(atlas_sce)%in%cells]
nr_Mixl1_positive <- table(chimera_Mixl1_tomato_pos$closest.cell[chimera_Mixl1_tomato_pos$closest.cell%in%colnames(atlas_sub)])
nr_Mixl1_negative <- table(chimeraMixl1_tomato_neg$closest.cell[chimeraMixl1_tomato_neg$closest.cell%in%colnames(atlas_sub)])
atlas_Mixl1_tomato_pos <- atlas_sub[,colnames(atlas_sub)%in%chimera_Mixl1_tomato_pos$closest.cell]
atlas_Mixl1_tomato_pos$nr_cells <- nr_Mixl1_positive[colnames(atlas_Mixl1_tomato_pos)]
atlas_Mixl1_tomato_neg <- atlas_sub[,colnames(atlas_sub)%in%chimeraMixl1_tomato_neg$closest.cell]
atlas_Mixl1_tomato_neg$nr_cells <- nr_Mixl1_negative[colnames(atlas_Mixl1_tomato_neg)]
sce_temp <- cbind(atlas_Mixl1_tomato_pos,atlas_Mixl1_tomato_neg)
sce_temp$mapping <- c( rep("Mixl1 tomato pos",ncol(atlas_Mixl1_tomato_pos)),
                      rep("Mixl1 tomato neg",ncol(atlas_Mixl1_tomato_neg)))
atlas_sub$mapping <- "atlas"
atlas_sub$nr_cells <- 1
sce_temp <- sce_temp[,sample(1:ncol(sce_temp))]
sce_temp_3 <- cbind(atlas_sce,sce_temp)
  
  
p3 <- ggplot(mapping=aes(x=reducedDims(sce_temp_3)$umap[,1],
                   y=reducedDims(sce_temp_3)$umap[,2],
                   color=sce_temp_3$mapping,size=sce_temp_3$nr_cells)) + theme_classic(base_size=10)+ggtitle("Mixl1")+geom_jitter()+
  scale_color_manual(values=c( "Mixl1 tomato neg"="darkblue","Mixl1 tomato pos"="red","atlas"="lightgrey"))+xlab("umap 1")+ylab("umap 2") + theme(legend.position = "none", legend.title=element_blank(),axis.text.x=element_blank(),
      axis.ticks.x=element_blank(),
      axis.text.y=element_blank(),
      axis.ticks.y=element_blank(),legend.text=element_text(size=12)) 
print(p3)
```

```{r mapping_Mixl1_chimeras_legend,dev='svg',dev='pdf'}
legend_mapping_Mixl1_chimeras <- get_legend(p3)
as_ggplot(legend_mapping_Mixl1_chimeras)
```

